### 2014-05-07 docs/paper.tex
latex format

\documentclass[12pt,journal,compsoc]{IEEEtran}
\providecommand{\PSforPDF}[1]{#1}
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
\title{Evolutionary Methods for Particle Swarm Movement}
\author{Benjamin~Bengfort,~Kevin~Harrison,~and~Philip~Kim}

% make the title area
\maketitle

\begin{abstract}
Abstract goes here.
The abstract goes here.
\end{abstract}

\section{Introduction}

Ever since Reynolds demonstrated that flocks of birds could be simulated as multi-agent systems, wherein each agent's motion is determined solely by local interactions, there has been interest in extending such systems to solve more general problems. Rodriguez, for example, added to each agent a goal-driven intelligence in the form of a finite state machine (FSM) that switched the agent between different sets of movement behaviors according to its local environment and current goal. He was able to demonstrate that a team of such agents was able to solve a resource locate-and-collect problem, and that a team programmed with flocking behaviors outperformed teams of agents that did not influence each other's movements.

In this article, we have attempted to extend Rodriguez's findings by applying evolutionary techniques to the agent control mechanisms. Rodriguez determined the structure of the FSM and the parameters defining each state through inspection and empirical techniques. We attempt to determine the optimal parameters by an evolutionary process that runs a population of randomly-generated FSM configurations through a simulated problem and then evolves the population through fitness-based selection and random mutation.

Our hypothesis was that we could evolve a configuration for the FSM that would be competitive to a configuration tuned by a human.

\section{Methods}

Because we were attempting to demonstrate to an evolutionary process could process an agent controller that was  competitive with the hand-tuned controller used by Rodriguez, our experimental setup was intentionally similar to his: two teams of homogenous agents, each with a designated "home" location, deployed in a two-dimensional world with periodic boundary conditions.

As in Rodriguez, an agent's velocity at each timestep was composed of some combination of the following vectors:

Cohesion: a vector pointing towards the average position of friendly agents in the neighborhood, with magnitude equal to zero when the agent's distance to the average neighbor is zero and increasing quadratically with distance until it is equal to Vmax when the distance to the average neighbor is r.

Alignment: a vector pointing in the same direction as the average velocity of friendly agents in the neighborhood, with magnitude equal to zero when the agent's distance to the neighbors' average position is zero and increasing quadratically with distance until it is equal to Vmax when the distance to the average neighbor is r.

Separation: a vector pointing away from the average position of friendly agents in the neighborhood, with magnitude equal to Vmax when the distance between the agent and the average position is zero, and decreasing quadratically with distance until it is zero when the distance to the average neighbor is r.

%Seeking, Homing, Mineral Cohesion: vectors pointing directly towards the targetted site with magnitude Vmax.

Clearance: a vector pointing in a direction orthogonal to the difference between the average neighbor position and the agent's position with magnitude Vmax.

Avoidance: a vector pointing away from an enemy agent in the neighborhood, with magnitude Vmax when distance to the enemy is zero and decreasing linearly with distance until it is zero when the distance is r. Unlike all the other components, this can be applied multiple times each timestep--once for every enemy in the neighborhood.

Also per Rodriguez, the FSM controlling each agent was always in one of four states: searching for new resources, moving to the last known resource deposit, carrying resources back to the home location, or guarding the home or a resource deposit. Each state was composed of some combination of the previously described velocity components, each characterized by a numerical weight and the radius and angle that defined the neighborhood.

Unlike Rodriguez, who prioritized each component and applied them in order, discarding any component that would have caused the velocity to exceed the maximum velocity, velocity in our simulation was a simple linear combination of the components; the maximum velocity constraint was only applied at the end to the resulting sum. Also unlike Rodriguez, agents in our simulation had "inertia" and started each timestep with the velocity of the previous timestep.

Transitions between states were triggered by conditions in each agent's local environment according to hard-coded rules. For example, an agent in the searching state that detected a mineral deposit within a 200-unit radius (in any direction) would push that deposit onto the top of its memory stack and transition to the seeking state (whcih is characterized by movement toward the top location on its stack).

One of Rodriguez's main findings was that it was advantageous for the agents to post "guards" on their homes and/or on any discovered resource deposits. Agents that guarded only the home performed best of all, though agents that guarded both the home and deposits still bested non-guarding agents. In order for guarding behavior to be feasible, however, agents must avoid agents on the enemy team. Since we were allowing evolution to determine the strength of this avoidance parameter, we expected that in the absence of a "natural" motivation for avoidance, the avoidance parameter would be selected down to zero, allowing the agents to ignore enemy guards.

Therefore we implemented a penalty for colliding with another agent. An agent within 10 units of an enemy agent was rendered unable to move for a number of timesteps equal to 180 minus the angle of incidence in degrees. For example, if an agent collided with an opposing agent at right angles, the agent struck in the side would be unable to move for 90 timesteps, whereas the agent struck in the front would be unable to move for 180 timesteps. Similarly, being "rear-ended" by an opposing agent had no penalty at all. The hope was that by assigning the "responsible" agent more of the penalty, we would create an incentive to avoid collision.

In order to allow for the selection of guarding behavior, the transition to the guarding state was controlled by two  evolvable parameters -- the "home guarding threshold" and the "deposit guarding threshold" -- that specified the number of friendly agents considered sufficient to guard the respective sites; a value of 0 disabled guarding behavior. This was the only instance in our simulation where a transition between states was controlled by an evolvable parameter.

At the beginning of each timestep, therefore, each agent is in a particular state. For each velocity component in its current state, it determines the number of friendly and enemy agents in the respective neighborhood and calculates the direction and magnitude of the component. These components are multiplied by the respective weight, added to the agent's velocity from the previous timestep, and the resulting velocity is subjected to the Vmax constraint. The agent's position is updated by adding the velocity to the agent's position in the previous timestep. The agent then determines whether it should change its state based on the predefined transition rules and its current environment.

\section{Experimental Procedure}

The computational experiment we ran consisted of pitting two simulated teams in competition with each other in a 3000 x 3000 world for a limited time. We restricted each team to 10 agents, vice 50 in the Rodriguez paper, and used 5 resource deposits instead of 8; both changes were to make the simulation more computationally tractable. The "red" team's behavior was fixed to parameters similar to those found empirically by Rodriguez to yield the best results. At the start of evolution we randomly generated 50 configurations to control the "black" team. Fitness was defined as the number of resources gathered by the black team after 10,000 timesteps.

After running each of the 50 configurations, we subjected the population to a round of evolution. The most fit individual from the current generation was always carried into the next generation, with the other 49 selected via tournaments of size 3. Each of the 49 was then subjected to mutation, with an independent 20\% chance of mutating each weight, each radius, and each alpha of every component. If selected for mutation, the trait was changed by a random value uniformly-distributed between -0.2 and 0.2 for weights, -20 and 20 for radii, and -20 and 20 for alphas. Additionally, the home guarding and deposit guarding threshold each had a 20\% of changing by plus or minus 1.

\section{Results}
results

\section{Discussion}
This work described herein represents only the first steps in the application of evolutionary techniques to this problem domain. By mutating and applying selective pressure to the movement component parameters, we were able to show that a machine could optimize the parameters better than a person. This was a form of optimization, but it did not result in novel behaviors. Indeed, since we hard-wired the transitions between the states (allowing for the guarding threshold to evolve) and allowed the evolver to mutate existing movement components in a given state but not to add them (i.e., we pre-defined which components could have non-zero weight for each state), we held constant the "meaning" of each state.

The next step would be to evolve the structure of the FSM itself; not only to allow the evolver to mix-in any component to any state, but to allow it to add new states, delete old states, and change the transitions between the states.

\begin{thebibliography}{1}

\bibitem{IEEEhowto:kopka}
H.~Kopka and P.~W. Daly, \emph{A Guide to {\LaTeX}}, 3rd~ed.\hskip 1em plus
  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.

\end{thebibliography}

\end{document}

### 2014-05-07 docs/paper.tex
More tex stuff

@@
 
-Ever since Reynolds demonstrated that flocks of birds could be simulated as multi-agent systems, wherein each agent's motion is determined solely by local interactions, there has been interest in extending such systems to solve more general problems.
+Ever since Reynolds demonstrated that flocks of birds could be simulated as multi-agent systems, wherein each agent's motion is determined solely by local interactions, there has been interest in extending such systems to solve more general problems \cite{reynolds1987flocks}.
 Rodriguez, for example, added to each agent a goal-driven intelligence in the form of a finite state machine (FSM) that switched the agent between different sets of movement behaviors according to its local environment and current goal.
@@
 
-\begin{thebibliography}{1}
-
-\bibitem{IEEEhowto:kopka}
-H.~Kopka and P.~W. Daly, \emph{A Guide to {\LaTeX}}, 3rd~ed.\hskip 1em plus
-  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-
-\end{thebibliography}
+\bibliographystyle{plain}
+\bibliography{paper}
 

### 2014-05-07 setup.py
Setup scripts

@@ -1,1 +1,49 @@
-raise NotImplementedError("Still need to prepare the setup tool.")
+#!/usr/bin/env python
+
+try:
+    from setuptools import setup
+    from setuptools import find_packages
+except ImportError:
+    raise ImportError("Could not import \"setuptools\". Please install the setuptools package.")
+
+
+packages = find_packages(where=".", exclude=('tests', 'bin', 'docs', 'fixtures', 'assets', 'conf', 'deploy'))
+
+requires = []
+
+with open('requirements.txt', 'r') as reqfile:
+    for line in reqfile:
+        requires.append(line.strip())
+
+classifiers = (
+    'Development Status :: 4 - Beta',
+    'Environment :: Console',
+    'Enivronment :: No Input/Output (Daemon)',
+    'Environment :: MacOS X',
+    'Framework :: Celery',
+    'Framework :: PyGame',
+    'Intended Audience :: Science/Research',
+    'License :: OSI Approved :: MIT License',
+    'Natural Language :: English',
+    'Operating System :: MacOS :: MacOS X',
+    'Operating System :: POSIX :: Linux',
+    'Programming Language :: Python :: 2.7',
+    'Topic :: Scientific/Engineering :: Artificial Life',
+    'Topic :: Scientific/Engineering :: Artificial Intelligence',
+)
+
+config = {
+    "name": "Swarm-Simulator",
+    "version": "1.1",
+    "description": "Implements a search and rescue simulation as well as an evolver.",
+    "author": "Benjamin Bengfort",
+    "author_email": "benjamin@bengfort.com",
+    "url": "https://github.com/mclumd/swarm-simulator",
+    "packages": packages,
+    "install_requires": requires,
+    "classifiers": classifiers,
+    "zip_safe": False,
+    "scripts": ['bin/runsim.py', 'bin/evolver.py'],
+}
+
+setup(**config)

### 2014-05-07 setup.py
more install updates

@@ -33,7 +33,7 @@ classifiers = (
 )
 
 config = {
-    "name": "Swarm-Simulator",
+    "name": "swarm-simulator",
     "version": "1.1",
     "description": "Implements a search and rescue simulation as well as an evolver.",
     "author": "Benjamin Bengfort",

### 2014-05-07 docs/movement.tex
more writing

\documentclass{article}
\setlength{\parskip}{\baselineskip}%
\setlength{\parindent}{0pt}%
\begin{document}

The following are the movement vectors that have been defined in code.

\textbf{Cohesion}

Neighbors: team in sight by $r$ and $\alpha$ if they are not GUARDING or STUNNED.

\[ \vec { v_{ c } } =v_{ max }\frac { \Delta  \vec { p }  }{ \left\|   \Delta \vec { p }  \right\|  } \left( \frac { \left\|  \Delta \vec {  p }  \right\|  }{ r }  \right) ^{ 2 } \]

\textbf{Alignment}

Neighbors: team in sight by $r$ and $\alpha$ if they are SEEKING or SPREADING

\[ \vec { v_{ al } } =v_{ max }{\left( \frac { \left\| \Delta \vec { p } \right\|  }{ r }  \right)}^2\frac {\Delta \vec{v}} { \left\| \Delta \vec{v} \right\| } \]

\textbf{Avoidance}

Neighbors: enemies in sight by $r$ and $\alpha$ if not STUNNED

\[ \vec { v_{ av } } =\sum _{i=0}^{n} {v_{ max } \left( \frac{r - \left\| \Delta \vec{p_e} \right\|} {r} \right) \frac { \Delta \vec { p_e }  }{ \left\| \Delta \vec { p_e }  \right\|  }}  \]

\textbf{Separation}

Neighbors: team in sight by $r$ and $\alpha$ no matter their state

\[ \vec { v_{ d } } =-v_{ max }\frac {\Delta \vec{p}} { \left\| \Delta \vec{p} \right\| }{\left( \frac { r - \left\| \Delta \vec { p } \right\|  }{ r }  \right)}^2 \]

\textbf{Seeking/Homing/Mineral Cohesion}

Target: Some saved target either from memory or found in sight

\[ \vec{v_h} = v_{max} \frac {\vec{p_t} - \vec p} { \left\| \vec{p_t} - \vec p \right\| } \]

\textbf{Clearance}

Neighbors: team in sight by $r$ and $\alpha$ if they re not GUARDING or STUNNED

\[ \vec{v_{cl}} = v_{max} \frac {\Delta \vec p_\perp } { \left\| \Delta \vec p_\perp  \right\| }  \]

\end{document}

### 2014-05-07 docs/paper.tex
more writing

@@
 \begin{document}
-\title{Evolutionary Methods for Particle Swarm Movement}
+\title{Evolutionary Design of Particles for Collectively Intelligent Problem Solving}
 \author{Benjamin~Bengfort,~Kevin~Harrison,~and~Philip~Kim}
@@
 \begin{abstract}
-Abstract goes here.
-The abstract goes here.
+
+Emergent, self-organizing flocks of particles can be implemented using only simple, local movement behavior rules.
+When particles are also improved with a limited working memory and goals, the local interactions of these top-down controllers leads to emergent global problem solving such as search and retrieve.
+We will show in this paper that the rules for both movement and goals can be evolved using Evolutionary Computing techiques on the dynamics parameters, as well as on the finite state machines that control behavior.
+The evolved particles perform as well as or better then those designed by humans.
+
 \end{abstract}
@@
 
-Ever since Reynolds demonstrated that flocks of birds could be simulated as multi-agent systems, wherein each agent's motion is determined solely by local interactions, there has been interest in extending such systems to solve more general problems \cite{reynolds1987flocks}.
-Rodriguez, for example, added to each agent a goal-driven intelligence in the form of a finite state machine (FSM) that switched the agent between different sets of movement behaviors according to its local environment and current goal.
+Ever since Reynolds demonstrated that flocks of birds could be simulated as multi-agent systems \cite{reynolds1987flocks}, there has been interest in improving flocks, schools, and swarms to higher levels of collective intelligence.
+The relatively simple nature of each agent; its motion or goals are determined solely by local interactions; leads to emergent global behavior.
+Decentralized control would mean optimal and economic solutions to problems in both algorithmic and physical domains.
+Furthermore, there has been interest in extending such systems to solve more general problems than traditional optimization \cite{kennedy1995particle}, modeling complex systems \cite{ottino2003complex,clerc2002particle} or cultural algorithms \cite{chung1996testbed}.
+
+This extension has primarily focused on the physical domain to improve multirobot team movement control as in \cite{balch1998behavior,ccelikkanat2010steering,hodgins1994robot}, however higher order problem solving is currently being explored.
+One large area of research is the deployment of mobile robots or sensors for area coverage, either to map terrain, or to establish mesh communication links.
+In \cite{cheng2009distributed}, flocking behavior was used to maximize coverage, while still maintaining collective behavior.
+Other tasks being explored include urban pursuit \cite{winder2004using} and robot herding of animals \cite{vaughan1998robot}.
+
+Many approaches have improved the simple local dynamics by adding minimal, cost effective top down components that also behave locally to produce emergent behavior.
+One very successful addition to flocks for problem solving is the addition of working memory \cite{winder2012role,hu2003particle}.
+Building on flocks with memory improvements, Rodriguez in \cite{rodriguez2004extending} added goal-driven intelligence in the form of a finite state machine (FSM) that switched the agent between different sets of movement behaviors according to its local environment and current goal.
 He was able to demonstrate that a team of such agents was able to solve a resource locate-and-collect problem, and that a team programmed with flocking behaviors outperformed teams of agents that did not influence each other's movements.
+
+AMAS theory for designing complex adaptive systems \cite{capera2003amas}
+Designing swarms \cite{mataric1993designing}
+
+Evolutionary programming with swarms \cite{wei2002swarm,miranda2005evolutionary}
 

### 2014-05-07 docs/paper.tex
more writing

@@
 
-AMAS theory for designing complex adaptive systems \cite{capera2003amas}
-Designing swarms \cite{mataric1993designing}
+These exciting results seem to indicate that in the future difficult environments might be economically explored or exploited by flocking robotic teams exhibiting local behavior for collective intelligence.
+For example in disaster areas such as fires or earthquakes, teams of flocking robots may be able to more effectively search and rescue victims.
+Landscapes such as other planetary bodies, the deep sea, or volcanos may similarly be accessible by such teams.
+However, in order to work in varied terrains and tasks specific swarm systems must be developed by humans to design both the bottom up movement behaviors and the top down controllers.
+Such design tasks have even led to formal design theory such as \cite{mataric1993designing,capera2003amas}.
+Although this theory might be applicable to a wide array of problems, a more flexibile approach could enable even more adaptive systems.
 
-Evolutionary programming with swarms \cite{wei2002swarm,miranda2005evolutionary}
+Familiar evolutionary computation techniques such as Genetic Programming (GP) \cite{koza1992genetic}, Evolutionary Strategies \cite{rechenberg1989evolution}, and Genetic Algorithms \cite{goldberg1988genetic} have already shown proof of the potential of computational design techniques for machine learning, optimization, and even engineering.
+Finite state machines have already undergone evolutionary processes to produce better results for particular tasks like automatic target detection \cite{benson2000evolving}.
+Evolving movement behaviors would be the equivalent of other real value optimization problems.
+Using these techniques it seems that evolving the individual particles in a swarm would be an effective technique of natural design and adaptation that mimics nature.
 
-In this article, we have attempted to extend Rodriguez's findings by applying evolutionary techniques to the agent control mechanisms.
+This is not to be confused with evolutionary technqiues that leverage swarms as in \cite{wei2002swarm,miranda2005evolutionary}.
+In this paper, we have attempted to extend Rodriguez's findings in \cite{rodriguez2004extending} by applying evolutionary techniques specifically to the agent control mechanisms.
 Rodriguez determined the structure of the FSM and the parameters defining each state through inspection and empirical techniques.
-We attempt to determine the optimal parameters by an evolutionary process that runs a population of randomly-generated FSM configurations through a simulated problem and then evolves the population through fitness-based selection and random mutation.
+We attempt to determine the optimal parameters by an evolutionary process that runs a population of randomly-generated FSM configurations through a simulated problem and then evolves the population through fitness-based selection, recombination, and mutation genetic operators.
 
-Our hypothesis was that we could evolve a configuration for the FSM that would be competitive to a configuration tuned by a human.
+We will show that a novel finite state machine for movement control can be evolved using an evolutionary programming technique that performs as well as or better that those configured by a top-down design approach.
+We will present both the simulation using the best human designed parameters, then an evolutionary computation to evolve the individual particles in the simulation.
 
@@
 
-Because we were attempting to demonstrate to an evolutionary process could process an agent controller that was  competitive with the hand-tuned controller used by Rodriguez, our experimental setup was intentionally similar to his: two teams of homogenous agents, each with a designated "home" location, deployed in a two-dimensional world with periodic boundary conditions.
+In order to best demonstrate that an evolutionary process can generate an agent controller that is competitive with a human designed one, we have created an experimental setup that is intentionally similar to the work done in \cite{rodriguez2004extending}.
+The task is for a team of simulated agents to search and gather resources that are spread across a large two-dimensional world with periodic boundary conditions, then return the resources to a designated home base.
+
+The individuals of the team operate as a flock- local rules determining movement and goals lead to emergent flocking behavior.
+Each individual has a limited sight range and angle, the ability to communicate with other, nearby agents and a limited working memory.
+Particle behavior is governed by a finite state machine that determines how each particle moves given its individual state and the environment around it.
+
+The task is competitive, as there are two teams in the environment competing for the resources and therefore the simulation is non-deterministic.
+Agents can mine resources not only from designated depots, but also steal from the home base of the other team.
+The winner of a simulation is determined as the team that has the most resources in their home base at the end of a predetermined number of time steps.
+
+Fitness is the result of the simulation.
+One team is fixed as the best human designed team; the other is evolved.
+In this section we will present the details of the simulation itself, then the details of evolution.
+
+\subsection{Simulation}
 
@@
 
+\subsection{Evolution}
+
 \section{Experimental Procedure}

### 2014-05-07 docs/paper.tex
equations

@@
 
-\section{Methods}
+\section{Methodology}
 
@@
 
-Fitness is the result of the simulation.
-One team is fixed as the best human designed team; the other is evolved.
-In this section we will present the details of the simulation itself, then the details of evolution.
+The evolutionary process uses this simulation to pit individuals created via genetic operations against the best designed human agent.
+The fitness of the evolved team is a function of the resources that they have collected vs.
+the resources the other team collected against them.
+The next generation of evolution is then dependent of the fitness of teams as run in the simulation.
+In the parlance of evolutionary computing, the genotype is the real valued vector that represents the finite state machine of the particle and the phenotype is the emergent behavior of the team as a whole as it participates in the simulation.
+
+In this section we will present the details of the simulation, as well as the details of the evolutionary computation mechanism that searches for the optimal, most economic particle in the simulation.
 
@@
 
-As in Rodriguez, an agent's velocity at each timestep was composed of some combination of the following vectors:
+The movement behavior of particles in teams is governed by local interactions of particles within a particlar neighborhood.
+The particle updates its velocity at each timestep based on what it obeserves in the world around it.
+As a whole, this leads to self-organizing, emergent behavior of the flock as a combined whole of its individuals.
+In particular, movement behaviors are updated with six velocity components: cohesion, alignment, separation, seeking, clearance, and avoidance.
+Together these components enable flocking behavior, minimize collisions, and strike a balance between the exploration of the world and the exploitation of discovered minerals.
 
-Cohesion: a vector pointing towards the average position of friendly agents in the neighborhood, with magnitude equal to zero when the agent's distance to the average neighbor is zero and increasing quadratically with distance until it is equal to Vmax when the distance to the average neighbor is r.
+%TODO: Discuss maximum velocity and other simulation parameters
+%TODO: Discuss how the velocity is computed given the velocity components
+%TODO: Dicsuss the finite state machine and how it is composed of behaviors
 
-Alignment: a vector pointing in the same direction as the average velocity of friendly agents in the neighborhood, with magnitude equal to zero when the agent's distance to the neighbors' average position is zero and increasing quadratically with distance until it is equal to Vmax when the distance to the average neighbor is r.
+It is important to discuss each individual velocity component in detail, as these components are part of the evolutionary mechanism that we have employed.
+Every component has several parameters: the radius - the site distance and alpha, the site angle which determine the neighborhood of the particular component.
+Additionally the weight and priority of the component are used to compute the final velocity of the particle.
 
-Separation: a vector pointing away from the average position of friendly agents in the neighborhood, with magnitude equal to Vmax when the distance between the agent and the average position is zero, and decreasing quadratically with distance until it is zero when the distance to the average neighbor is r.
+Cohesion: a vector pointing towards the average position of friendly agents in the neighborhood, with magnitude equal to zero when the agent's distance to the average neighbor is zero and increasing quadratically with distance until it is equal to the maximum velocity when the distance to the average neighbor is r.
+The neighborhood includes all team members who are not guarding or stunned.
 
-%Seeking, Homing, Mineral Cohesion: vectors pointing directly towards the targetted site with magnitude Vmax.
+\[ \vec { v_{ c } } =v_{ max }\frac { \Delta  \vec { p }  }{ \left\|   \Delta \vec { p }  \right\|  } \left( \frac { \left\|  \Delta \vec {  p }  \right\|  }{ r }  \right) ^{ 2 } \]
 
-Clearance: a vector pointing in a direction orthogonal to the difference between the average neighbor position and the agent's position with magnitude Vmax.
+Alignment: a vector pointing in the same direction as the average velocity of friendly agents in the neighborhood, with magnitude equal to zero when the agent's distance to the neighbors' average position is zero and increasing quadratically with distance until it is equal to the maximum velocity when the distance to the average neighbor is r.
+The neighborhood includes all team members if they are seeking or spreading.
+
+\[ \vec { v_{ al } } =v_{ max }{\left( \frac { \left\| \Delta \vec { p } \right\|  }{ r }  \right)}^2\frac {\Delta \vec{v}} { \left\| \Delta \vec{v} \right\| } \]
+
+Separation: a vector pointing away from the average position of friendly agents in the neighborhood, with magnitude equal to the maximum velocity when the distance between the agent and the average position is zero, and decreasing quadratically with distance until it is zero when the distance to the average neighbor is r.
+The nighborhood includes all team members no matter their state.
+
+\[ \vec { v_{ d } } =-v_{ max }\frac {\Delta \vec{p}} { \left\| \Delta \vec{p} \right\| }{\left( \frac { r - \left\| \Delta \vec { p } \right\|  }{ r }  \right)}^2 \]
+
+Seeking, Homing, and Mineral Cohesion: vectors pointing directly towards the target site with a magnitude proportional to the maximum velocity.
+These velocity components allow the particle to navigate ina  meaningful manner.
+The target is stored in memory and this velocity component isn't influenced by nearby neighbors.
+
+\[ \vec{v_h} = v_{max} \frac {\vec{p_t} - \vec p} { \left\| \vec{p_t} - \vec p \right\| } \]
+
+Clearance: a vector pointing in a direction orthogonal to the difference between the average neighbor position and the agent's position proportional the the maximum velocity.
+This component ensures that swarms have a wide field of view.
+The neighborhood of this component is team members in sight that are not guarding or stunned.
+
+\[ \vec{v_{cl}} = v_{max} \frac {\Delta \vec p_\perp } { \left\| \Delta \vec p_\perp  \right\| }  \]
 
@@
 Unlike all the other components, this can be applied multiple times each timestep--once for every enemy in the neighborhood.
+
+\[ \vec { v_{ av } } =\sum _{i=0}^{n} {v_{ max } \left( \frac{r - \left\| \Delta \vec{p_e} \right\|} {r} \right) \frac { \Delta \vec { p_e }  }{ \left\| \Delta \vec { p_e }  \right\|  }}  \]
 

### 2014-05-08 docs/paper.tex
experiment

@@
 \hyphenation{op-tical net-works semi-conduc-tor}
+\usepackage{graphicx}
+\usepackage{float}
+\floatstyle{boxed}
+\restylefloat{figure}
 
@@
 
-\section{Experimental Procedure}
-
-The computational experiment we ran consisted of pitting two simulated teams in competition with each other in a 3000 x 3000 world for a limited time.
-We restricted each team to 10 agents, vice 50 in the Rodriguez paper, and used 5 resource deposits instead of 8; both changes were to make the simulation more computationally tractable.
-The "red" team's behavior was fixed to parameters similar to those found empirically by Rodriguez to yield the best results.
-At the start of evolution we randomly generated 50 configurations to control the "black" team.
-Fitness was defined as the number of resources gathered by the black team after 10,000 timesteps.
-
 After running each of the 50 configurations, we subjected the population to a round of evolution.
@@
 Additionally, the home guarding and deposit guarding threshold each had a 20\% of changing by plus or minus 1.
+
+\section{Experimental Procedure}
+
+Our computational experiment required two main components; a system to carry out the evolutionary computation, which constructed genotype populations via genetic operators, and a simulation system to instantiate phenotypes- particle swarms whose individuals are constructed from the genotypes.
+The simulation system computed the fitness of a particular swarm, which allowed the evolutionary computation to build the next generation.
+
+The simulation environment consisted of two simulated teams in competition with each other; a red team and a black team.
+The simulated world was a 3000 x 3000 world with periodic boundary conditions and the number of timesteps was limited to 10,000 per simulation due to computational constraints.
+We restricted each team to 10 agents each,and evenly distributed 5 resource deposits betwen the red and black bases with 80 resources per deposit.
+Both the team size and the time limit were implemented to make the evolutionary computation tractable; our implementation required on average 273 seconds to complete; and each generation took approximately 52 minutes to compute all fitness values.
+
+The red team's behavior was implemented using a finite state machine, memory, and movement parameters exactly as the best empirical result designed by humans in \cite{rodriguez2004extending}, called \textit{full guarding flock} meaning that the swarm guarded both its home base as well as discovered mineral deposits.
+The black team utilized the configuration derived from a genotype individual produced by the evolutionary computation.
+We attempted to hold as many variables constant as possible, including the locations of bases and resources, starting positions and velocities of agents, and update order.
+After 10,000 timesteps, the fitness of the black team was returned as the number of resources in the black base.
+
+\begin{figure}[h!]
+    \centering
+        \includegraphics[width=0.5\textwidth]{figures/simulation}
+    \caption{Simulation of agents spreading in flocks.}
+\end{figure}
+
+Each simulation was run in parallel as subprocess of a master evolutionary computation process.
+The master process randomly initialized a population of 50 black team configurations, then ran a simulation for each individual in the population.
+After all 50 simulations were completed, tournament based selection with tournaments of size three were carried over to the next generation along with the 5 most elite individuals.
+After selection, genetic operators were applied.
+
+In our first experiment we only used mutation to change the states in the finite state machine of the particles.
+In the second experiment we used recombination in addition to mutation operations.
+In either case, elite individuals were exempt from genetic operations, however elite individuals could be recombined to replace other individuals in the population.
+Both the mutation and recombination (when used) had a likelihood of 0.2.
+Maximum and minimum parameters were also placed on the values of the weight, the radii, and the alpha values to ensure an economic and reasonable result.
 

### 2014-05-08 docs/paper.tex
Start evolution subsection of paper

--- 2014-05-07 docs/paper.tex
@@
 
+The evolution of a finite state machine for movement control was achieved through the application of a simple set of genetic operators to the weights and parameters of the machine.
+The genotype of a movement controller was represented as two integers, one for each guarding threshold, and a vector of real-valued numbers consisting of the alpha parameter, radius, and weight given to each component of each compound movement behavior.
+
+The set of states in each controller was fixed to be the same set of compound behaviors as those found in \cite{rodriguez2004extending}, but with parameters derived through evolutionary programming.
+This allowed for easy comparison with the authoritative, human-derived solution.
+In addition, the transitions between states were kept the same except for the guarding thresholds, which were evolved.
+
+New individuals were introduced through the application of mutation and recombination operators.
+The mutation operator adds or removes an integer value from each selected threshold and adds a uniformly distributed random value with mean 0 to each selected real-valued parameter.
+Each mutation operation selects which parameters to modify independently, so the mutation of one parameter is independent of the mutation of another parameter.
+
 \section{Experimental Procedure}

### 2014-05-08 docs/paper.tex
In the middle of revising abstract and intro.

--- 2014-05-07 docs/paper.tex
@@
 
-Emergent, self-organizing flocks of particles can be implemented using only simple, local movement behavior rules.
-When particles are also improved with a limited working memory and goals, the local interactions of these top-down controllers leads to emergent global problem solving such as search and retrieve.
-We will show in this paper that the rules for both movement and goals can be evolved using Evolutionary Computing techiques on the dynamics parameters, as well as on the finite state machines that control behavior.
-The evolved particles perform as well as or better then those designed by humans.
+Self-organizing "flocks" of agents that display emergent coordinated behavior can be implemented using only local interaction rules.
+When agents are also given memory and a goal-based Finite State Machine (FSM) controller, the resulting team is able to solve a locate-and-collect task.
+We show that Evolutionary Computing techiques can be applied to the parameters that comprise the FSM states.
+Comparing the resulting FSMs to similar controllers described in the previous literature, we find that the evolved FSMs perform as well as or better than those designed by human  observation.
+We close by suggesting methods for applying evolutionary techniques to the structure as well as to the numerical parameters of the controller, perhaps leading to the emergence of novel behavior.
 
@@
 
-Ever since Reynolds demonstrated that flocks of birds could be simulated as multi-agent systems \cite{reynolds1987flocks}, there has been interest in improving flocks, schools, and swarms to higher levels of collective intelligence.
-The relatively simple nature of each agent; its motion or goals are determined solely by local interactions; leads to emergent global behavior.
-Decentralized control would mean optimal and economic solutions to problems in both algorithmic and physical domains.
-Furthermore, there has been interest in extending such systems to solve more general problems than traditional optimization \cite{kennedy1995particle}, modeling complex systems \cite{ottino2003complex,clerc2002particle} or cultural algorithms \cite{chung1996testbed}.
+Ever since Reynolds demonstrated that flocks of birds could be simulated as multi-agent systems \cite{reynolds1987flocks}, wherein each agent's motion is determined solely by local interactions with other agents, there has been interest in extending such systems to solve more general problems.
+The self-organizing characteristic of these systems, whereby coordinated behavior emerges without a central control mechanism, suggests that even relatively simple decentralized systems might yield novel approaches to problems in both the computational and physical domains.
 
-This extension has primarily focused on the physical domain to improve multirobot team movement control as in \cite{balch1998behavior,ccelikkanat2010steering,hodgins1994robot}, however higher order problem solving is currently being explored.
-One large area of research is the deployment of mobile robots or sensors for area coverage, either to map terrain, or to establish mesh communication links.
-In \cite{cheng2009distributed}, flocking behavior was used to maximize coverage, while still maintaining collective behavior.
+By abstracting the idea of "flocks" to $n$-dimensional spaces, for example, "particle swarms" \cite{kennedy1995particle,clerc2002particle} or "cultural algorithms" \cite{chung1996testbed} have been used for numerical optimization.
+In the physical domain, swarm behavior have been used multirobot team movement control \cite{balch1998behavior,ccelikkanat2010steering,hodgins1994robot}, or during deployment of mobile robots or sensors.
+In \cite{cheng2009distributed}, flocking behavior was used to maximize coverage during such deployments while still maintaining collective behavior.
 Other tasks being explored include urban pursuit \cite{winder2004using} and robot herding of animals \cite{vaughan1998robot}.
 
-Many approaches have improved the simple local dynamics by adding minimal, cost effective top down components that also behave locally to produce emergent behavior.
-One very successful addition to flocks for problem solving is the addition of working memory \cite{winder2012role,hu2003particle}.
-Building on flocks with memory improvements, Rodriguez in \cite{rodriguez2004extending} added goal-driven intelligence in the form of a finite state machine (FSM) that switched the agent between different sets of movement behaviors according to its local environment and current goal.
+Many approaches augment the agents with such features as working memory \cite{winder2012role,hu2003particle}.
+Rodriguez \cite{rodriguez2004extending} added both memory and a finite state machine (FSM) controller that switched the agent between different sets of movement behaviors according to its local environment and current goal.
 He was able to demonstrate that a team of such agents was able to solve a resource locate-and-collect problem, and that a team programmed with flocking behaviors outperformed teams of agents that did not influence each other's movements.
 
-These exciting results seem to indicate that in the future difficult environments might be economically explored or exploited by flocking robotic teams exhibiting local behavior for collective intelligence.
-For example in disaster areas such as fires or earthquakes, teams of flocking robots may be able to more effectively search and rescue victims.
-Landscapes such as other planetary bodies, the deep sea, or volcanos may similarly be accessible by such teams.
-However, in order to work in varied terrains and tasks specific swarm systems must be developed by humans to design both the bottom up movement behaviors and the top down controllers.
-Such design tasks have even led to formal design theory such as \cite{mataric1993designing,capera2003amas}.
-Although this theory might be applicable to a wide array of problems, a more flexibile approach could enable even more adaptive systems.
+Attempts have been made to formalize the design of such cooperative multi-agent systems \cite{mataric1993designing,capera2003amas}.
+In this paper, however, we turn to evolutionary computation techniques.
+Genetic Programming (GP) \cite{koza1992genetic}, Evolutionary Strategies \cite{rechenberg1989evolution}, and Genetic Algorithms \cite{goldberg1988genetic} have demonstrated the potential of computational design techniques for machine learning, optimization, and even engineering.
+Finite state machines have been evolved to solve problems such as automatic target detection \cite{benson2000evolving}.
+Evolving the FSM described by Rodriguez, therefore, seems like a natural extension of multi-agent problem-solving.
+This is not to be confused with evolutionary techniques that leverage swarms as in \cite{wei2002swarm,miranda2005evolutionary}.
 
-Familiar evolutionary computation techniques such as Genetic Programming (GP) \cite{koza1992genetic}, Evolutionary Strategies \cite{rechenberg1989evolution}, and Genetic Algorithms \cite{goldberg1988genetic} have already shown proof of the potential of computational design techniques for machine learning, optimization, and even engineering.
-Finite state machines have already undergone evolutionary processes to produce better results for particular tasks like automatic target detection \cite{benson2000evolving}.
-Evolving movement behaviors would be the equivalent of other real value optimization problems.
-Using these techniques it seems that evolving the individual particles in a swarm would be an effective technique of natural design and adaptation that mimics nature.
-
-This is not to be confused with evolutionary technqiues that leverage swarms as in \cite{wei2002swarm,miranda2005evolutionary}.
 In this paper, we have attempted to extend Rodriguez's findings in \cite{rodriguez2004extending} by applying evolutionary techniques specifically to the agent control mechanisms.
@@
 
-%TODO: Discuss maximum velocity and other simulation parameters
-%TODO: Discuss how the velocity is computed given the velocity components
-%TODO: Dicsuss the finite state machine and how it is composed of behaviors
-
 It is important to discuss each individual velocity component in detail, as these components are part of the evolutionary mechanism that we have employed.
@@
 
-Avoidance: a vector pointing away from an enemy agent in the neighborhood, with magnitude Vmax when distance to the enemy is zero and decreasing linearly with distance until it is zero when the distance is r.
+Avoidance: a vector pointing away from an enemy agent in the neighborhood, with magnitude $v_{max}$ when distance to the enemy is zero and decreasing linearly with distance until it is zero when the distance is r.
 Unlike all the other components, this can be applied multiple times each timestep--once for every enemy in the neighborhood.
@@
 For each velocity component in its current state, it determines the number of friendly and enemy agents in the respective neighborhood and calculates the direction and magnitude of the component.
-These components are multiplied by the respective weight, added to the agent's velocity from the previous timestep, and the resulting velocity is subjected to the Vmax constraint.
+These components are multiplied by the respective weight, added to the agent's velocity from the previous timestep, and the resulting velocity is subjected to the $v_{max}$ constraint.
 The agent's position is updated by adding the velocity to the agent's position in the previous timestep.

### 2014-05-08 docs/paper.tex
Finished abstract/intro revisions.

@@
 Finite state machines have been evolved to solve problems such as automatic target detection \cite{benson2000evolving}.
-Evolving the FSM described by Rodriguez, therefore, seems like a natural extension of multi-agent problem-solving.
-This is not to be confused with evolutionary techniques that leverage swarms as in \cite{wei2002swarm,miranda2005evolutionary}.
+Evolving the FSM described by Rodriguez (not to be confused with evolutionary techniques that leverage swarms as in \cite{wei2002swarm,miranda2005evolutionary}), therefore, seems like a natural extension of multi-agent problem-solving.
 
-In this paper, we have attempted to extend Rodriguez's findings in \cite{rodriguez2004extending} by applying evolutionary techniques specifically to the agent control mechanisms.
-Rodriguez determined the structure of the FSM and the parameters defining each state through inspection and empirical techniques.
-We attempt to determine the optimal parameters by an evolutionary process that runs a population of randomly-generated FSM configurations through a simulated problem and then evolves the population through fitness-based selection, recombination, and mutation genetic operators.
-
-We will show that a novel finite state machine for movement control can be evolved using an evolutionary programming technique that performs as well as or better that those configured by a top-down design approach.
-We will present both the simulation using the best human designed parameters, then an evolutionary computation to evolve the individual particles in the simulation.
+In this paper, we have attempted to extend Rodriguez's findings by applying evolutionary techniques to the agent control mechanisms.
+Rodriguez determined the structure of the FSM and the parameters defining each state through empirical techniques and inspection.
+In contrast, we begin with a population of randomly-generated FSM configurations, run them through a simulated problem, and then evolve the population over many generations through fitness-based selection, recombination, and mutation.
+Our hypothesis is that the resulting FSM configuration performs as well as or better than a configuration tuned by a human.
 

### 2014-05-08 docs/paper.tex

--- 2014-05-08 docs/paper.tex
@@
 
-Self-organizing "flocks" of agents that display emergent coordinated behavior can be implemented using only local interaction rules.
-When agents are also given memory and a goal-based Finite State Machine (FSM) controller, the resulting team is able to solve a locate-and-collect task.
-We show that Evolutionary Computing techiques can be applied to the parameters that comprise the FSM states.
-Comparing the resulting FSMs to similar controllers described in the previous literature, we find that the evolved FSMs perform as well as or better than those designed by human  observation.
-We close by suggesting methods for applying evolutionary techniques to the structure as well as to the numerical parameters of the controller, perhaps leading to the emergence of novel behavior.
+Emergent, self-organizing flocks of particles can be implemented using only simple, local movement behavior rules.
+When particles are also improved with a limited working memory and goals, the local interactions of these top-down controllers leads to emergent global problem solving such as search and retrieve.
+We will show in this paper that the rules for both movement and goals can be evolved using Evolutionary Computing techiques on the dynamics parameters, as well as on the finite state machines that control behavior.
+The evolved particles perform as well as or better then those designed by humans.
 
@@
 
-Ever since Reynolds demonstrated that flocks of birds could be simulated as multi-agent systems \cite{reynolds1987flocks}, wherein each agent's motion is determined solely by local interactions with other agents, there has been interest in extending such systems to solve more general problems.
-The self-organizing characteristic of these systems, whereby coordinated behavior emerges without a central control mechanism, suggests that even relatively simple decentralized systems might yield novel approaches to problems in both the computational and physical domains.
+Ever since Reynolds demonstrated that flocks of birds could be simulated as multi-agent systems \cite{reynolds1987flocks}, there has been interest in improving flocks, schools, and swarms to higher levels of collective intelligence.
+The relatively simple nature of each agent; its motion or goals are determined solely by local interactions; leads to emergent global behavior.
+Decentralized control would mean optimal and economic solutions to problems in both algorithmic and physical domains.
+Furthermore, there has been interest in extending such systems to solve more general problems than traditional optimization \cite{kennedy1995particle}, modeling complex systems \cite{ottino2003complex,clerc2002particle} or cultural algorithms \cite{chung1996testbed}.
 
-By abstracting the idea of "flocks" to $n$-dimensional spaces, for example, "particle swarms" \cite{kennedy1995particle,clerc2002particle} or "cultural algorithms" \cite{chung1996testbed} have been used for numerical optimization.
-In the physical domain, swarm behavior have been used multirobot team movement control \cite{balch1998behavior,ccelikkanat2010steering,hodgins1994robot}, or during deployment of mobile robots or sensors.
-In \cite{cheng2009distributed}, flocking behavior was used to maximize coverage during such deployments while still maintaining collective behavior.
+This extension has primarily focused on the physical domain to improve multirobot team movement control as in \cite{balch1998behavior,ccelikkanat2010steering,hodgins1994robot}, however higher order problem solving is currently being explored.
+One large area of research is the deployment of mobile robots or sensors for area coverage, either to map terrain, or to establish mesh communication links.
+In \cite{cheng2009distributed}, flocking behavior was used to maximize coverage, while still maintaining collective behavior.
 Other tasks being explored include urban pursuit \cite{winder2004using} and robot herding of animals \cite{vaughan1998robot}.
 
-Many approaches augment the agents with such features as working memory \cite{winder2012role,hu2003particle}.
-Rodriguez \cite{rodriguez2004extending} added both memory and a finite state machine (FSM) controller that switched the agent between different sets of movement behaviors according to its local environment and current goal.
+Many approaches have improved the simple local dynamics by adding minimal, cost effective top down components that also behave locally to produce emergent behavior.
+One very successful addition to flocks for problem solving is the addition of working memory \cite{winder2012role,hu2003particle}.
+Building on flocks with memory improvements, Rodriguez in \cite{rodriguez2004extending} added goal-driven intelligence in the form of a finite state machine (FSM) that switched the agent between different sets of movement behaviors according to its local environment and current goal.
 He was able to demonstrate that a team of such agents was able to solve a resource locate-and-collect problem, and that a team programmed with flocking behaviors outperformed teams of agents that did not influence each other's movements.
 
-Attempts have been made to formalize the design of such cooperative multi-agent systems \cite{mataric1993designing,capera2003amas}.
-In this paper, however, we turn to evolutionary computation techniques.
-Genetic Programming (GP) \cite{koza1992genetic}, Evolutionary Strategies \cite{rechenberg1989evolution}, and Genetic Algorithms \cite{goldberg1988genetic} have demonstrated the potential of computational design techniques for machine learning, optimization, and even engineering.
-Finite state machines have been evolved to solve problems such as automatic target detection \cite{benson2000evolving}.
-Evolving the FSM described by Rodriguez, therefore, seems like a natural extension of multi-agent problem-solving.
-This is not to be confused with evolutionary techniques that leverage swarms as in \cite{wei2002swarm,miranda2005evolutionary}.
+These exciting results seem to indicate that in the future difficult environments might be economically explored or exploited by flocking robotic teams exhibiting local behavior for collective intelligence.
+For example in disaster areas such as fires or earthquakes, teams of flocking robots may be able to more effectively search and rescue victims.
+Landscapes such as other planetary bodies, the deep sea, or volcanos may similarly be accessible by such teams.
+However, in order to work in varied terrains and tasks specific swarm systems must be developed by humans to design both the bottom up movement behaviors and the top down controllers.
+Such design tasks have even led to formal design theory such as \cite{mataric1993designing,capera2003amas}.
+Although this theory might be applicable to a wide array of problems, a more flexibile approach could enable even more adaptive systems.
 
+Familiar evolutionary computation techniques such as Genetic Programming (GP) \cite{koza1992genetic}, Evolutionary Strategies \cite{rechenberg1989evolution}, and Genetic Algorithms \cite{goldberg1988genetic} have already shown proof of the potential of computational design techniques for machine learning, optimization, and even engineering.
+Finite state machines have already undergone evolutionary processes to produce better results for particular tasks like automatic target detection \cite{benson2000evolving}.
+Evolving movement behaviors would be the equivalent of other real value optimization problems.
+Using these techniques it seems that evolving the individual particles in a swarm would be an effective technique of natural design and adaptation that mimics nature.
+
+This is not to be confused with evolutionary technqiues that leverage swarms as in \cite{wei2002swarm,miranda2005evolutionary}.
 In this paper, we have attempted to extend Rodriguez's findings in \cite{rodriguez2004extending} by applying evolutionary techniques specifically to the agent control mechanisms.
@@
 
+%TODO: Discuss maximum velocity and other simulation parameters
+%TODO: Discuss how the velocity is computed given the velocity components
+%TODO: Dicsuss the finite state machine and how it is composed of behaviors
+
 It is important to discuss each individual velocity component in detail, as these components are part of the evolutionary mechanism that we have employed.
@@
 
-Avoidance: a vector pointing away from an enemy agent in the neighborhood, with magnitude $v_{max}$ when distance to the enemy is zero and decreasing linearly with distance until it is zero when the distance is r.
+Avoidance: a vector pointing away from an enemy agent in the neighborhood, with magnitude Vmax when distance to the enemy is zero and decreasing linearly with distance until it is zero when the distance is r.
 Unlike all the other components, this can be applied multiple times each timestep--once for every enemy in the neighborhood.
@@
 For each velocity component in its current state, it determines the number of friendly and enemy agents in the respective neighborhood and calculates the direction and magnitude of the component.
-These components are multiplied by the respective weight, added to the agent's velocity from the previous timestep, and the resulting velocity is subjected to the $v_{max}$ constraint.
+These components are multiplied by the respective weight, added to the agent's velocity from the previous timestep, and the resulting velocity is subjected to the Vmax constraint.
 The agent's position is updated by adding the velocity to the agent's position in the previous timestep.
@@
 
-After running each of the 50 configurations, we subjected the population to a round of evolution.
-The most fit individual from the current generation was always carried into the next generation, with the other 49 selected via tournaments of size 3.
-Each of the 49 was then subjected to mutation, with an independent 20\% chance of mutating each weight, each radius, and each alpha of every component.
-If selected for mutation, the trait was changed by a random value uniformly-distributed between -0.2 and 0.2 for weights, -20 and 20 for radii, and -20 and 20 for alphas.
-Additionally, the home guarding and deposit guarding threshold each had a 20\% of changing by plus or minus 1.
+The evolution of a finite state machine for movement control was achieved through the application of a simple set of genetic operators to the weights and parameters of the machine.
+The genotype of a movement controller was represented as two integers, one for each guarding threshold, and a vector of real-valued numbers consisting of the alpha parameter, radius, and weight given to each component of each compound movement behavior.
+
+The set of states in each controller was fixed to be the same set of compound behaviors as those found in \cite{rodriguez2004extending}, but with parameters derived through evolutionary programming.
+This allowed for easy comparison with the authoritative, human-derived solution.
+In addition, the transitions between states were kept the same except for the guarding thresholds, which we evolved.
+New individuals were introduced through the application of mutation and recombination operators.
+
+The mutation operator first selected which parameters to modify independently.
+We used a 20\% chance for mutating each weight, radius, alpha, and threshold.
+Each selected threshold had an equal chance of increasing by 1 and decreasing by 1.
+Each selected real-valued parameter was modified through the addition of a uniformly distributed random number with mean 0.
+We chose to mutate each weight by -0.2 to 0.2 and each alpha and radius by -20 to 20.
+
+The recombination operator ... %TODO talk about recombination
+
+A round of evolution combined the mutation and recombination operators to produce the next generation.
+After evaluating every individual in the current generation, we applied elitism, tournament selection, mutation, and recombination to complete a round of evolution.
+For elitism, the single best individual of the previous generation was always carried into the next generation without modification.
+The other 49 individuals were selected via tournaments of size 3.
+Each of the 49 was then subjected to the mutation operator, with its parameters being mutated independently.
 

### 2014-05-08 docs/paper.tex
results figs

--- 2014-05-08 docs/paper.tex
@@
 Finite state machines have been evolved to solve problems such as automatic target detection \cite{benson2000evolving}.
-Evolving the FSM described by Rodriguez (not to be confused with evolutionary techniques that leverage swarms as in \cite{wei2002swarm,miranda2005evolutionary}), therefore, seems like a natural extension of multi-agent problem-solving.
+Evolving the FSM described by Rodriguez, therefore, seems like a natural extension of multi-agent problem-solving.
+This is not to be confused with evolutionary techniques that leverage swarms as in \cite{wei2002swarm,miranda2005evolutionary}.
 
-In this paper, we have attempted to extend Rodriguez's findings by applying evolutionary techniques to the agent control mechanisms.
-Rodriguez determined the structure of the FSM and the parameters defining each state through empirical techniques and inspection.
-In contrast, we begin with a population of randomly-generated FSM configurations, run them through a simulated problem, and then evolve the population over many generations through fitness-based selection, recombination, and mutation.
-Our hypothesis is that the resulting FSM configuration performs as well as or better than a configuration tuned by a human.
+In this paper, we have attempted to extend Rodriguez's findings in \cite{rodriguez2004extending} by applying evolutionary techniques specifically to the agent control mechanisms.
+Rodriguez determined the structure of the FSM and the parameters defining each state through inspection and empirical techniques.
+We attempt to determine the optimal parameters by an evolutionary process that runs a population of randomly-generated FSM configurations through a simulated problem and then evolves the population through fitness-based selection, recombination, and mutation genetic operators.
+
+We will show that a novel finite state machine for movement control can be evolved using an evolutionary programming technique that performs as well as or better that those configured by a top-down design approach.
+We will present both the simulation using the best human designed parameters, then an evolutionary computation to evolve the individual particles in the simulation.
 
@@
 \section{Results}
-results
+
+\begin{figure}[h!]
+    \centering
+        \includegraphics[width=0.5\textwidth]{figures/meanfit_experiment1}
+    \caption{Mean fitness per generation with mutation}
+\end{figure}
 

### 2014-05-08 docs/paper.tex
results

--- 2014-05-08 docs/paper.tex
@@
 \hyphenation{op-tical net-works semi-conduc-tor}
+\usepackage[font=small,labelfont=bf]{caption}
 \usepackage{graphicx}
@@
 
-% make the title area
-\maketitle
-
+\IEEEcompsoctitleabstractindextext{%
 \begin{abstract}
@@
 
-\end{abstract}
+\end{abstract}}
+
+% make the title area
+\maketitle
+
+\IEEEdisplaynotcompsoctitleabstractindextext
 
@@
 \section{Results}
-results
+
+In this section we will evaluate the effectiveness of both evolved flocking particles vs.
+human created particles, as well as the ability of an evolutionary computation to design such particles.
+In short, the evolutionary computation mechanism was able to design a particle system that exhibited variations in behaviors that were different enough from the human designed particles so as not only be competitive with the human design team, but able to defeat the other team in terms of resources collected.
+
+Table \ref{table:meanfitness} shows the mean, standard deviation, maximum, and minimum resources collected for the black team over ten runs of the simulation with the same parameters.
+Black teams of comparsion are the human designed team, the best evolved team, and a random team.
+The evolution designed team performs significantly better than the human designed team.
+
+\begin{table}
+    \renewcommand{\arraystretch}{1.5}
+    \centering
+    \begin{tabular}{l | c c cc}
+        \hline
+        Simulation & Mean & Std Dev & Max & Min \\
+        \hline
+        human & 119.0 & 43.2 & 274 & 119 \\
+        evolved & 296.0 & 25.6 & 336 & 254 \\
+        random & 0.3 & 0.67 & 2 & 0 \\
+        \hline
+    \end{tabular}
+    \caption{Aggregated fitness of various particles competing with human designed particles.}
+    \label{table:meanfitness}
+\end{table}
+
+The average fitness over time of the black and red teams is displayed in \ref{fig:head2head} which graphs the mean fitness of ten individual runs of the best human designed team (the red team) vs the best evolved team, the black team.
+Within 6000 timesteps the red team begins to lose resources as the black team exploits the red base in addition to more resource locations than the red team, which favors a strategy of exploiting one resource deposit at a time.
+
+\begin{figure}[h!]
+    \centering
+        \includegraphics[width=0.5\textwidth]{figures/head2head}
+    \caption{Performance of evolved particles vs human designed particles in head to head simulation of 10,000 timesteps}
+    \label{fig:head2head}
+\end{figure}
+
+The performance of our two evolutionary computations is expressed in \ref{fig:meanfit1} and \ref{fig:meanfit2} which plots the mean fitness of the entire population per each generation.
+In experiment one, only the mutation operator was used to generate the next population.
+In experiment two, both recombination and mutation were used to produce the next generation.
+
+\begin{figure}[h!]
+    \centering
+        \includegraphics[width=0.5\textwidth]{figures/meanfit_experiment1}
+    \caption{Mean fitness per generation with mutation}
+    \label{fig:meanfit1}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+        \includegraphics[width=0.5\textwidth]{figures/fitness21}
+    \caption{Mean fitness per generation with mutation and recombination}
+    \label{fig:meanfit2}
+\end{figure}
+
+As is shown from the graphs; the first experiment was wildly unstable in terms of its search across the optimal state space.
+This may have been due to the fact that highly variable mutation caused dramatic shifts in the behavior of the phenotypes.
+It also may have been due to a programming error on the part of the research team.
+In the second evolution with recombination as well as mutation, there is more steadily growing fitness over time and a more stable population to select from to apply genetic operations.
 
 \section{Discussion}
+The evolution designed particles surpassed their human design counterparts through a qualitatively novel mechanism; the human designed team prefered to stay in one or two larger flocks; whereas the high performing evolved teams split into more, smaller flocks and were able to exploit the search space more effectively.
+
 This work described herein represents only the first steps in the application of evolutionary techniques to this problem domain.

### 2014-05-08 bin/runsim.py

@@ -21,6 +21,7 @@ Run the swarm simulation.
 
 import os
 import sys
+import csv
 import time
 import argparse
 import cProfile
@@ -100,6 +101,35 @@ def profile(args):
     cProfile.runctx('run()', globals(), locals(), args.filename, args.sort)
     return ''
 
+def head2head(args):
+    """
+    Run a head to head simulation, outputing the magnitude of the stash of
+    each team in the simulation at every time step.
+    """
+    start = time.time()
+    world = World(ally_conf_path=args.conf_path, maximum_time=args.iterations)
+
+    print "Starting headless simulation, use CTRL+C to quit."
+    writer = csv.writer(args.stream, delimiter='\t')
+    writer.writerow(('black', 'red'))
+    while world.time < world.iterations:
+        try:
+            world.update()
+            writer.writerow((str(world.ally_home.stash), str(world.enemy_home.stash)))
+            if world.time % 1000 == 0:
+                print "%ik iterations completed" % (world.time / 1000)
+        except KeyboardInterrupt:
+            print "Quitting Early!"
+            break
+
+    finit = time.time()
+    delta = finit - start
+
+    output = []
+    output.append("Ran %i time steps in %0.3f seconds" % (world.time, delta))
+    output.append("Agents successfully collected %i resources" % world.ally_home.stash)
+    return "\n".join(output)
+
 ##########################################################################
 ## Main method
 ##########################################################################
@@ -133,6 +163,14 @@ def main(*argv):
                                 help='Number of iterations to profile')
     profile_parser.add_argument('-c', '--conf-path', type=str, dest='conf_path', default='./conf/params.yaml', help='path to ally configuration file.')
     profile_parser.set_defaults(func=profile)
+
+    # head2head headless simulation
+    head2head_parser = subparsers.add_parser('head2head', help='Run a headless simulation with the configuration file')
+    head2head_parser.add_argument('-c', '--conf-path', type=str, dest='conf_path', default='./conf/params.yaml', help='path to ally configuration file.')
+    head2head_parser.add_argument('-o', '--outpath', dest='stream', type=argparse.FileType('w'), default=sys.stdout, help='Write head to head results out.')
+    head2head_parser.add_argument('-i', '--iterations', metavar='STEPS', type=int, default=10000,
+                                    help='Number of iterations to profile')
+    head2head_parser.set_defaults(func=head2head)
 
     # Handle input from the command line
     args = parser.parse_args()            # Parse the arguments

### 2014-05-08 docs/paper.tex
Describe recombination

@@
 This allowed for easy comparison with the authoritative, human-derived solution.
-In addition, the transitions between states were kept the same except for the guarding thresholds, which we evolved.
+In addition, the transitions between states were kept the same except for the guarding thresholds, which were evolved.
 New individuals were introduced through the application of mutation and recombination operators.
@@
 Each selected real-valued parameter was modified through the addition of a uniformly distributed random number with mean 0.
-We chose to mutate each weight by -0.2 to 0.2 and each alpha and radius by -20 to 20.
+We mutated each weight by -0.2 to 0.2 and each alpha and radius by -20 to 20.
 
-The recombination operator ... %TODO talk about recombination
+The recombination operator replaces one individual with a combination of it and another individual in the population.
+We used intermediary recombination of the weight, radius, and alpha parameters of all components in all states, bounded by the minimum and maximum possible values of each parameter.
 
-A round of evolution combined the mutation and recombination operators to produce the next generation.
-After evaluating every individual in the current generation, we applied elitism, tournament selection, mutation, and recombination to complete a round of evolution.
+A round of evolution consisted of the evaluation of every individual followed by the application of elitism, tournament selection, recombination, and mutation to produce the next generation.
 For elitism, the single best individual of the previous generation was always carried into the next generation without modification.
 The other 49 individuals were selected via tournaments of size 3.
-Each of the 49 was then subjected to the mutation operator, with its parameters being mutated independently.
+The 49 individuals chosen through the tournament then each had a 20\% chance of undergoing recombination with another randomly chosen child.
+Each of the resulting 49 was then subjected to the mutation operator, mutating parameters independently.
 

### 2014-05-08 docs/paper.tex
moving to editing mode

--- 2014-05-08 docs/paper.tex
@@
 
-\[ \vec { v_{ c } } =v_{ max }\frac { \Delta  \vec { p }  }{ \left\|   \Delta \vec { p }  \right\|  } \left( \frac { \left\|  \Delta \vec {  p }  \right\|  }{ r }  \right) ^{ 2 } \]
+\begin{equation}
+    \vec { v_{ c } } =v_{ max }\frac { \Delta  \vec { p }  }{ \left\|   \Delta \vec { p }  \right\|  } \left( \frac { \left\|  \Delta \vec {  p }  \right\|  }{ r }  \right) ^{ 2 }
+    \label{eqn:cohesion}
+\end{equation}
 
@@
 
-\[ \vec { v_{ al } } =v_{ max }{\left( \frac { \left\| \Delta \vec { p } \right\|  }{ r }  \right)}^2\frac {\Delta \vec{v}} { \left\| \Delta \vec{v} \right\| } \]
+\begin{equation}
+    \vec { v_{ al } } =v_{ max }{\left( \frac { \left\| \Delta \vec { p } \right\|  }{ r }  \right)}^2\frac {\Delta \vec{v}} { \left\| \Delta \vec{v} \right\| }
+    \label{eqn:alignment}
+\end{equation}
 
@@
 
-\[ \vec { v_{ d } } =-v_{ max }\frac {\Delta \vec{p}} { \left\| \Delta \vec{p} \right\| }{\left( \frac { r - \left\| \Delta \vec { p } \right\|  }{ r }  \right)}^2 \]
+\begin{equation}
+    \vec { v_{ d } } =-v_{ max }\frac {\Delta \vec{p}} { \left\| \Delta \vec{p} \right\| }{\left( \frac { r - \left\| \Delta \vec { p } \right\|  }{ r }  \right)}^2
+    \label{eqn:separation}
+\end{equation}
 
@@
 
-\[ \vec{v_h} = v_{max} \frac {\vec{p_t} - \vec p} { \left\| \vec{p_t} - \vec p \right\| } \]
+\begin{equation}
+    \vec{v_h} = v_{max} \frac {\vec{p_t} - \vec p} { \left\| \vec{p_t} - \vec p \right\| }
+    \label{eqn:seeking}
+\end{equation}
 
@@
 
-\[ \vec{v_{cl}} = v_{max} \frac {\Delta \vec p_\perp } { \left\| \Delta \vec p_\perp  \right\| }  \]
+\begin{equation}
+    \vec{v_{cl}} = v_{max} \frac {\Delta \vec p_\perp } { \left\| \Delta \vec p_\perp  \right\| }
+    \label{eqn:clearance}
+\end{equation}
 
@@
 
-\[ \vec { v_{ av } } =\sum _{i=0}^{n} {v_{ max } \left( \frac{r - \left\| \Delta \vec{p_e} \right\|} {r} \right) \frac { \Delta \vec { p_e }  }{ \left\| \Delta \vec { p_e }  \right\|  }}  \]
+\begin{equation}
+    \vec { v_{ av } } =\sum _{i=0}^{n} {v_{ max } \left( \frac{r - \left\| \Delta \vec{p_e} \right\|} {r} \right) \frac { \Delta \vec { p_e }  }{ \left\| \Delta \vec { p_e }  \right\|  }}
+    \label{eqn:avoidance}
+\end{equation}
 
@@
 
-The recombination operator ... %TODO talk about recombination
+The recombination operator also ignored the elite individuals carried over from the last population.
+Parents were selected with a likelihood of 0.2 to be recombined with another individual.
+Our recombination operator replaced the first parent with intermediary recombination, that is the mean value of the parents velocity parameters.
+The second chosen parent remained in the population.
+Although elite individuals were exempt from recombination, they were able to be selected as the secondary parent, contributing their genetic material to the recombined offspring.
 
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \centering
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \centering
-        \includegraphics[width=0.5\textwidth]{figures/fitness21}
+        \includegraphics[width=0.5\textwidth]{figures/fitness46}
     \caption{Mean fitness per generation with mutation and recombination}

### 2014-05-08 docs/paper.tex

@@
 Each selected real-valued parameter was modified through the addition of a uniformly distributed random number with mean 0.
-We chose to mutate each weight by -0.2 to 0.2 and each alpha and radius by -20 to 20.
+We mutated each weight by -0.2 to 0.2 and each alpha and radius by -20 to 20.
 
@@
 Parents were selected with a likelihood of 0.2 to be recombined with another individual.
-Our recombination operator replaced the first parent with intermediary recombination, that is the mean value of the parents velocity parameters.
+We used intermediary recombination of the weight, radius, and alpha parameters of all components in all states, bounded by the minimum and maximum possible values of each parameter.
 The second chosen parent remained in the population.
@@
 
-A round of evolution combined the mutation and recombination operators to produce the next generation.
-After evaluating every individual in the current generation, we applied elitism, tournament selection, mutation, and recombination to complete a round of evolution.
+A round of evolution consisted of the evaluation of every individual followed by the application of elitism, tournament selection, recombination, and mutation to produce the next generation.
 For elitism, the single best individual of the previous generation was always carried into the next generation without modification.
 The other 49 individuals were selected via tournaments of size 3.
-Each of the 49 was then subjected to the mutation operator, with its parameters being mutated independently.
+The 49 individuals chosen through the tournament then each had a 20\% chance of undergoing recombination with another randomly chosen child.
+Each of the resulting 49 was then subjected to the mutation operator, mutating parameters independently.
 

### 2014-05-08 docs/paper.tex
all but discussion

@@
 \begin{document}
-\title{Evolutionary Design of Particles for Collectively Intelligent Problem Solving}
+\title{Evolutionary Design of Flocking Particles for Collectively Intelligent Problem Solving}
 \author{Benjamin~Bengfort,~Kevin~Harrison,~and~Philip~Kim}
@@
 When agents are also given memory and a goal-based Finite State Machine (FSM) controller, the resulting team is able to solve a locate-and-collect task.
-We show that Evolutionary Computing techiques can be applied to the parameters that comprise the FSM states.
+We show that Evolutionary Computing techiques can be applied to the FSM controller, where the genotype is a real valued vector representing the FSM and the phenotype is the emergent, self-organizing behavior.
 Comparing the resulting FSMs to similar controllers described in the previous literature, we find that the evolved FSMs perform as well as or better than those designed by human  observation.
-We close by suggesting methods for applying evolutionary techniques to the structure as well as to the numerical parameters of the controller, perhaps leading to the emergence of novel behavior.
+In this paper we will propose methods for applying evolutionary techniques to the structure and states of the controller, perhaps leading to the emergence of novel behavior.
 
@@
 
-By abstracting the idea of "flocks" to $n$-dimensional spaces, for example, "particle swarms" \cite{kennedy1995particle,clerc2002particle} or "cultural algorithms" \cite{chung1996testbed} have been used for numerical optimization.
-In the physical domain, swarm behavior have been used multirobot team movement control \cite{balch1998behavior,ccelikkanat2010steering,hodgins1994robot}, or during deployment of mobile robots or sensors.
-In \cite{cheng2009distributed}, flocking behavior was used to maximize coverage during such deployments while still maintaining collective behavior.
-Other tasks being explored include urban pursuit \cite{winder2004using} and robot herding of animals \cite{vaughan1998robot}.
+By abstracting the idea of "flocks" to $n$-dimensional spaces, for example, particle swarms \cite{kennedy1995particle,clerc2002particle} or cultural algorithms \cite{chung1996testbed} have been used for numerical optimization.
+In the physical domain, swarm behaviors have been used to control multirobot team movement \cite{balch1998behavior,ccelikkanat2010steering,hodgins1994robot}, or during deployment of mobile robots or sensors.
+In \cite{cheng2009distributed}, flocking behavior was used to maximize area coverage during such deployments coordinating movements via collective behavior.
+Other tasks using collective problem solving being explored include urban pursuit \cite{winder2004using} and robot herding of animals \cite{vaughan1998robot}.
 
 Many approaches augment the agents with such features as working memory \cite{winder2012role,hu2003particle}.
-Rodriguez \cite{rodriguez2004extending} added both memory and a finite state machine (FSM) controller that switched the agent between different sets of movement behaviors according to its local environment and current goal.
-He was able to demonstrate that a team of such agents was able to solve a resource locate-and-collect problem, and that a team programmed with flocking behaviors outperformed teams of agents that did not influence each other's movements.
+Our paper builds upon the work of Rodriguez & Reggia in \cite{rodriguez2004extending}, who added both working memory and a finite state machine (FSM) controller that switched the agent between different sets of movement behaviors according to its local environment and current goal.
+They were able to demonstrate that a team of such agents was able to solve a resource locate-and-collect problem, and that a team programmed with flocking behaviors outperformed teams of agents that did not influence each other's movements.
 
@@
 Genetic Programming (GP) \cite{koza1992genetic}, Evolutionary Strategies \cite{rechenberg1989evolution}, and Genetic Algorithms \cite{goldberg1988genetic} have demonstrated the potential of computational design techniques for machine learning, optimization, and even engineering.
+Swarm techniques have even been applied to evolutionary computation itself as in  \cite{wei2002swarm,miranda2005evolutionary}.
+However, we are more interested in the opposite application, using evolutionary techniques to determine swarm behavior.
 Finite state machines have been evolved to solve problems such as automatic target detection \cite{benson2000evolving}.
-Evolving the FSM described by Rodriguez (not to be confused with evolutionary techniques that leverage swarms as in \cite{wei2002swarm,miranda2005evolutionary}), therefore, seems like a natural extension of multi-agent problem-solving.
+Evolving the high level controllers therefore, seems like an extension of the design of multi-agent problem-solving that is inspired by natural design.
 
-In this paper, we have attempted to extend Rodriguez's findings by applying evolutionary techniques to the agent control mechanisms.
-Rodriguez determined the structure of the FSM and the parameters defining each state through empirical techniques and inspection.
-In contrast, we begin with a population of randomly-generated FSM configurations, run them through a simulated problem, and then evolve the population over many generations through fitness-based selection, recombination, and mutation.
-Our hypothesis is that the resulting FSM configuration performs as well as or better than a configuration tuned by a human.
+In this paper, we have improved the human design of particles, particularly those described in \cite{rodriguez2004extending} by applying evolutionary techniques to the agent control mechanisms.
+A formal human design technique determines the structure of the FSM and the parameters defining each state through empirical techniques and inspection.
+In contrast, we begin with a population of randomly-generated FSM configurations, run them through a simulated problem, then evolve the population over many generations through fitness-based selection, recombination, and mutation.
+Our hypothesis is that the resulting FSM configuration will perform as well as or better than a configuration tuned by a human.
 
@@
 In order to best demonstrate that an evolutionary process can generate an agent controller that is competitive with a human designed one, we have created an experimental setup that is intentionally similar to the work done in \cite{rodriguez2004extending}.
-The task is for a team of simulated agents to search and gather resources that are spread across a large two-dimensional world with periodic boundary conditions, then return the resources to a designated home base.
+The task is for a team of simulated agents to locate and collect resources that are spread across a large two-dimensional world with periodic boundary conditions, then return the resources to a designated home base.
 
-The individuals of the team operate as a flock- local rules determining movement and goals lead to emergent flocking behavior.
+The individuals of the team operate as a flock- local rules determining individual movement and goals lead to emergent collective behavior.
 Each individual has a limited sight range and angle, the ability to communicate with other, nearby agents and a limited working memory.
@@
 the resources the other team collected against them.
-The next generation of evolution is then dependent of the fitness of teams as run in the simulation.
+The next generation of evolution is then dependent on the fitness of teams as run in the simulation.
 In the parlance of evolutionary computing, the genotype is the real valued vector that represents the finite state machine of the particle and the phenotype is the emergent behavior of the team as a whole as it participates in the simulation.
@@
 
-The movement behavior of particles in teams is governed by local interactions of particles within a particlar neighborhood.
+The movement behavior of individual particles is governed by local interactions of particles within a particlar neighborhood.
 The particle updates its velocity at each timestep based on what it obeserves in the world around it.
-As a whole, this leads to self-organizing, emergent behavior of the flock as a combined whole of its individuals.
+This leads to self-organizing, emergent behavior of the flock as a combined whole of its individuals.
 In particular, movement behaviors are updated with six velocity components: cohesion, alignment, separation, seeking, clearance, and avoidance.
@@
 
-It is important to discuss each individual velocity component in detail, as these components are part of the evolutionary mechanism that we have employed.
-Every component has several parameters: the radius - the site distance and alpha, the site angle which determine the neighborhood of the particular component.
-Additionally the weight and priority of the component are used to compute the final velocity of the particle.
+It is important to discuss each individual velocity component in detail, as these components define the movement behaviors of the states that compose the FSM controller that is being evolved.
+Every component has several parameters: a radius, $r$ and alpha, $\alpha$, which determine the sight distance and angle of the neighborhood of the particular component.
+Each component additionaly has a weight, $w$, which is used to determine the contribution of the component to the final velocity of the particle.
 
-Cohesion: a vector pointing towards the average position of friendly agents in the neighborhood, with magnitude equal to zero when the agent's distance to the average neighbor is zero and increasing quadratically with distance until it is equal to the maximum velocity when the distance to the average neighbor is r.
+\subsubsection{Velocity Components}
+
+All particles are bound by a maximum velocity, $v_{max}$ and for the purposes of evolution, a maximum radius, $r_{max}$ to simulate economic constraints in the world.
+Other notation used is as follows: $\Delta \vec {p} = \vec {p_n} - \vec {p}$ where $\vec {p_n}$ is the average of the neighbor positions and $\vec p$ is the position of the agent.
+Similarly $\Delta \vec {v} = \vec {v_n} - \vec v$ where $\vec {v_n}$ is the average of the neighbor velocities and $\vec v$ is the velocity of the agent.
+Other positions include $\vec p_e$ which is the position of an enemy agent and $\vec p_t$, the position of a target and $\vec p_\perp$ represents a unit vector ortogonal to the agent in the $+z$ direction.
+Specific velocity components are defined as follows.
+
+\textit{Cohesion} (\ref{eqn:cohesion}): a vector pointing towards the average position of friendly agents in the neighborhood, with a magnitude proportional to the agent's distance to the center of its neighbors.
+This component increases quadratically with distance until it is equal to the maximum velocity when the distance to the neighbor center is $r$.
 The neighborhood includes all team members who are not guarding or stunned.
+This velocity component ensures that flocks form and stay together throughout movement.
 
@@
 
-Alignment: a vector pointing in the same direction as the average velocity of friendly agents in the neighborhood, with magnitude equal to zero when the agent's distance to the neighbors' average position is zero and increasing quadratically with distance until it is equal to the maximum velocity when the distance to the average neighbor is r.
+\textit{Alignment} (\ref{eqn:alignment}): a vector in the same direction as the average velocity of friendly agents in the neighborhood, with magnitude proportional to the agent's distance to the center of its neighbors.
+This component increases quadratically with distance until it is equal to the maximum velocity when the distance neighbor center is $r$.
 The neighborhood includes all team members if they are seeking or spreading.
+Alignment helps the flock travel in the same direction in a coordinated fashion.
 
@@
 
-Separation: a vector pointing away from the average position of friendly agents in the neighborhood, with magnitude equal to the maximum velocity when the distance between the agent and the average position is zero, and decreasing quadratically with distance until it is zero when the distance to the average neighbor is r.
+\textit{Separation} (\ref{eqn:separation}): a vector pointing away from the average position of friendly agents in the neighborhood, with magnitude inversely proportional to the distance from the agent to the center of its neighbors.
+This component decreases quadratically with distance until it is zero when the distance to the neighbor center is $r$.
 The nighborhood includes all team members no matter their state.
+Separation prevents particles in the same team from colliding with each other, as well as distributing the flock outwards.
 
@@
 
-Seeking, Homing, and Mineral Cohesion: vectors pointing directly towards the target site with a magnitude proportional to the maximum velocity.
-These velocity components allow the particle to navigate ina  meaningful manner.
+\textit{Seeking, Homing, and Mineral Cohesion} (\ref{eqn:seeking}): vectors pointing directly towards a target with a magnitude proportional to the maximum velocity.
+These velocity components allow the particle to navigate in a  meaningful manner.
 The target is stored in memory and this velocity component isn't influenced by nearby neighbors.
@@
 
-Clearance: a vector pointing in a direction orthogonal to the difference between the average neighbor position and the agent's position proportional the the maximum velocity.
-This component ensures that swarms have a wide field of view.
+\textit{Clearance} (\ref{eqn:clearance}): a vector pointing in a direction orthogonal to the difference between the average neighbor position and the agent's position proportional the the maximum velocity.
+This component ensures that swarms have a wide field of view by spreading the flock laterally along the front of movement.
 The neighborhood of this component is team members in sight that are not guarding or stunned.
@@
 
-Avoidance: a vector pointing away from an enemy agent in the neighborhood, with magnitude $v_{max}$ when distance to the enemy is zero and decreasing linearly with distance until it is zero when the distance is r.
+\textit{Avoidance} (\ref{eqn:avoidance}): a vector pointing away from an enemy agent in the neighborhood, with magnitude $v_{max}$ when distance to the enemy is zero and decreasing linearly with distance until it is zero when the distance is $r$.
 Unlike all the other components, this can be applied multiple times each timestep--once for every enemy in the neighborhood.
@@
 
-Also per Rodriguez, the FSM controlling each agent was always in one of four states: searching for new resources, moving to the last known resource deposit, carrying resources back to the home location, or guarding the home or a resource deposit.
-Each state was composed of some combination of the previously described velocity components, each characterized by a numerical weight and the radius and angle that defined the neighborhood.
+\subsubsection{Movement Behavior}
 
-Unlike Rodriguez, who prioritized each component and applied them in order, discarding any component that would have caused the velocity to exceed the maximum velocity, velocity in our simulation was a simple linear combination of the components; the maximum velocity constraint was only applied at the end to the resulting sum.
-Also unlike Rodriguez, agents in our simulation had "inertia" and started each timestep with the velocity of the previous timestep.
+The FSM controlling each agent in our first experiment contained four states: searching for new resources (SPREADING), moving to the last known resource deposit (SEEKING), carrying resources back to the home location (CARAVANING), or guarding the home base or a resource deposit (GUARDING).
+Each state was composed of some combination of the previously described velocity components.
 
-Transitions between states were triggered by conditions in each agent's local environment according to hard-coded rules.
-For example, an agent in the searching state that detected a mineral deposit within a 200-unit radius (in any direction) would push that deposit onto the top of its memory stack and transition to the seeking state (whcih is characterized by movement toward the top location on its stack).
+To update the velocity of a particle at time $t$, a simple linear combination of velocity components was used.
+The components, along with their parameters and weights depended on the current state of the particle.
+The magnitude (or length) of the vector was maximized at $v_{max}$ in the direction of the computed component velocity.
+Our simulation also implemented "inertia", whereby the velocity was not combined from zero, but rather added to the velocity at time $t-1$.
+The position of the particle at time $t$ was simply the sum of the position at time $t-1$ with the new velocity.
 
-One of Rodriguez's main findings was that it was advantageous for the agents to post "guards" on their homes and/or on any discovered resource deposits.
-Agents that guarded only the home performed best of all, though agents that guarded both the home and deposits still bested non-guarding agents.
+\begin{equation}
+    \vec {v_t} =  \vec {v_{t-1}}+ \sum_i^c { w_i v_i}
+    \label{eqn:velocity}
+\end{equation}
+
+\begin{equation}
+    \vec {p_t} = \vec {p_{t-1}} + \vec {v_t}
+    \label{eqn:position}
+\end{equation}
+
+Transitions between states were triggered by conditions in each agent's local environment according to rules in the FSM (in our first experiment, these rules were fixed).
+For example, an agent in the searching state that detected a mineral deposit within a 200-unit radius (in any direction) would push that deposit onto the top of its memory stack and transition to the seeking state (which is characterized by movement toward the top location on its stack).
+
+One of the main findings in \cite{rodriguez2004extending} was that it was advantageous for the agents to post "guards" on their homes and/or on any discovered resource deposits.
+Agents that guarded only the home performed best of all, though agents that guarded both the home and deposits still outperformed non-guarding agents.
 In order for guarding behavior to be feasible, however, agents must avoid agents on the enemy team.
-Since we were allowing evolution to determine the strength of this avoidance parameter, we expected that in the absence of a "natural" motivation for avoidance, the avoidance parameter would be selected down to zero, allowing the agents to ignore enemy guards.
+Since we were allowing evolution to determine the strength of this avoidance parameter, we expected that in the absence of an enforced motivation for avoidance, the avoidance parameter would be selected down to zero, allowing the agents to ignore enemy guards.
 
@@
 
-In order to allow for the selection of guarding behavior, the transition to the guarding state was controlled by two  evolvable parameters -- the "home guarding threshold" and the "deposit guarding threshold" -- that specified the number of friendly agents considered sufficient to guard the respective sites; a value of 0 disabled guarding behavior.
-This was the only instance in our simulation where a transition between states was controlled by an evolvable parameter.
+\begin{figure}[h!]
+    \centering
+        \includegraphics[width=0.5\textwidth]{figures/simulation}
+    \caption{Simulation of agents spreading in flocks.}
+\end{figure}
 
-At the beginning of each timestep, therefore, each agent is in a particular state.
+In order to allow for the selection of guarding behavior, the transition to the guarding state was controlled by two  evolvable parameters -- the \textit{home guarding threshold} and the \textit{deposit guarding threshold}, which specified the number of friendly agents considered sufficient to guard the respective sites; a value of 0 disabled guarding behavior.
+This was the only instance in our simulation where a transition between states was controlled by an evolvable parameter, and we will explore adding this component in future work.
+
+A single timestep in the world can be characterized as follows.
+Each agent is in a particular state.
 For each velocity component in its current state, it determines the number of friendly and enemy agents in the respective neighborhood and calculates the direction and magnitude of the component.
-These components are multiplied by the respective weight, added to the agent's velocity from the previous timestep, and the resulting velocity is subjected to the $v_{max}$ constraint.
+These components are linearly summed, and the resulting velocity is subjected to the $v_{max}$ constraint.
 The agent's position is updated by adding the velocity to the agent's position in the previous timestep.
 The agent then determines whether it should change its state based on the predefined transition rules and its current environment.
+All agents make this computation simultaneously, updating does not affect the world in the current time step.
 
@@
 
-The evolution of a finite state machine for movement control was achieved through the application of a simple set of genetic operators to the weights and parameters of the machine.
-The genotype of a movement controller was represented as two integers, one for each guarding threshold, and a vector of real-valued numbers consisting of the alpha parameter, radius, and weight given to each component of each compound movement behavior.
+The evolution of a finite state machine for movement control was achieved through the application of a simple set of genetic operators to the states of the FSM as defined by a collection of velocity components and their behavior specific parameters.
+The genotype for our simulation was represented as a vector containing two integers, one for each guarding threshold, and real-valued numbers consisting of the alpha parameter, radius, and weight given to each component of each compound movement behavior.
 
-The set of states in each controller was fixed to be the same set of compound behaviors as those found in \cite{rodriguez2004extending}, but with parameters derived through evolutionary programming.
+In this experiment, the set of states in each controller was fixed to be the same set of compound behaviors as those found in \cite{rodriguez2004extending}, but with parameters derived through evolutionary programming.
 This allowed for easy comparison with the authoritative, human-derived solution.
 In addition, the transitions between states were kept the same except for the guarding thresholds, which were evolved.
-New individuals were introduced through the application of mutation and recombination operators.
+The basic evolutionary computation was a mixture of Evolutionary Strategy and Evolutionary Programming techniques and is characterized in Table \ref{table:evolvemethod}.
 
 The mutation operator first selected which parameters to modify independently.
-We used a 20\% chance for mutating each weight, radius, alpha, and threshold.
+We used a mutation probability parameter $p_m=0.2$  for mutating each weight, radius, alpha, and threshold.
 Each selected threshold had an equal chance of increasing by 1 and decreasing by 1.
 Each selected real-valued parameter was modified through the addition of a uniformly distributed random number with mean 0.
-We mutated each weight by -0.2 to 0.2 and each alpha and radius by -20 to 20.
+We mutated each weight with world constraints using $r\in\left(1,500\right]$ and $\alpha\in\left(0,360\right]$, the step size of the mutation was $m_w\in\left[-0.2,0.2\right]$ for weights and $m_s\in\left[20,20\right]$ for the $r$ and $\alpha$ parameters.
+Elite individuals were exempted from mutation.
 
 The recombination operator also ignored the elite individuals carried over from the last population.
-Parents were selected with a likelihood of 0.2 to be recombined with another individual.
+Parents were selected with a recombination probability, $p_r=0.2$, to be recombined with another individual.
 We used intermediary recombination of the weight, radius, and alpha parameters of all components in all states, bounded by the minimum and maximum possible values of each parameter.
@@
 A round of evolution consisted of the evaluation of every individual followed by the application of elitism, tournament selection, recombination, and mutation to produce the next generation.
-For elitism, the single best individual of the previous generation was always carried into the next generation without modification.
-The other 49 individuals were selected via tournaments of size 3.
-The 49 individuals chosen through the tournament then each had a 20\% chance of undergoing recombination with another randomly chosen child.
-Each of the resulting 49 was then subjected to the mutation operator, mutating parameters independently.
+For elitism, the $N_e$ individuals of the previous generation was always carried into the next generation without modification.
+The other $p-N_e$ individuals were selected via tournaments of size 3.
+All non-elite individuals chosen through the tournament then each had a $p_r$ chance of undergoing recombination with another randomly chosen child.
+Each of the resulting non-elite individuals was then subjected to the mutation operator, mutating parameters independently.
+
+\begin{table}
+    \renewcommand{\arraystretch}{1.5}
+    \centering
+    \begin{tabular}{|l|l|}
+        \hline
+        Genotype & Real valued vectors representing FSM states \\
+        Phenotype & Swarm behavior from individual movements \\
+        Selection & Tournament with Elitism \\
+        Recombination & Intermediary exempting elites \\
+        Mutation & Linear Rule exempting elites \\
+        Fitness & Resources collected after 10k timesteps \\
+        Specialty & Adpating FSM with same size network \\
+        \hline
+    \end{tabular}
+    \caption{Characterization of Evolutionary Method}
+    \label{table:evolvemethod}
+\end{table}
 
@@
 
-The red team's behavior was implemented using a finite state machine, memory, and movement parameters exactly as the best empirical result designed by humans in \cite{rodriguez2004extending}, called \textit{full guarding flock} meaning that the swarm guarded both its home base as well as discovered mineral deposits.
+The red team's behavior was implemented using a finite state machine, memory, and movement parameters exactly as the best empirical result designed by humans in \cite{rodriguez2004extending}, called \textit{home guarding flock} meaning that the swarm guarded both its home base but not discovered mineral deposits.
 The black team utilized the configuration derived from a genotype individual produced by the evolutionary computation.
@@
 After 10,000 timesteps, the fitness of the black team was returned as the number of resources in the black base.
-
-\begin{figure}[h!]
-    \centering
-        \includegraphics[width=0.5\textwidth]{figures/simulation}
-    \caption{Simulation of agents spreading in flocks.}
-\end{figure}
 
@@
 
-Table \ref{table:meanfitness} shows the mean, standard deviation, maximum, and minimum resources collected for the black team over ten runs of the simulation with the same parameters.
-Black teams of comparsion are the human designed team, the best evolved team, and a random team.
-The evolution designed team performs significantly better than the human designed team.
-
 \begin{table}
@@
 
-The average fitness over time of the black and red teams is displayed in \ref{fig:head2head} which graphs the mean fitness of ten individual runs of the best human designed team (the red team) vs the best evolved team, the black team.
-Within 6000 timesteps the red team begins to lose resources as the black team exploits the red base in addition to more resource locations than the red team, which favors a strategy of exploiting one resource deposit at a time.
-
 \begin{figure}[ht!]
@@
 
-The performance of our two evolutionary computations is expressed in \ref{fig:meanfit1} and \ref{fig:meanfit2} which plots the mean fitness of the entire population per each generation.
+Table \ref{table:meanfitness} shows the mean, standard deviation, maximum, and minimum resources collected for the black team over ten runs of the simulation with the same parameters.
+Black teams of comparsion are the human designed team, the best evolved team, and a random team.
+The evolution designed team performs significantly better than the human designed team.
+
+The average fitness over time of the black and red teams is displayed in Figure \ref{fig:head2head} which graphs the mean fitness of ten individual runs of the best human designed team (the red team) vs the best evolved team, the black team.
+Within 6000 timesteps the red team begins to lose resources as the black team exploits the red base in addition to more resource locations than the red team, which favors a strategy of exploiting one resource deposit at a time.
+
+The performance of our two evolutionary computations is expressed in Figures \ref{fig:meanfit1} and \ref{fig:meanfit2} which plots the mean fitness of the entire population per each generation.
 In experiment one, only the mutation operator was used to generate the next population.
 In experiment two, both recombination and mutation were used to produce the next generation.
+
+As is shown from the graphs; the first experiment was wildly unstable in terms of its search across the optimal state space.
+This may have been due to the fact that highly variable mutation caused dramatic shifts in the behavior of the phenotypes.
+It also may have been due to a programming error on the part of the research team.
+In the second evolution with recombination as well as mutation, there is more steadily growing fitness over time and a more stable population to select from to apply genetic operations.
 
@@
 
-As is shown from the graphs; the first experiment was wildly unstable in terms of its search across the optimal state space.
-This may have been due to the fact that highly variable mutation caused dramatic shifts in the behavior of the phenotypes.
-It also may have been due to a programming error on the part of the research team.
-In the second evolution with recombination as well as mutation, there is more steadily growing fitness over time and a more stable population to select from to apply genetic operations.
-
 \section{Discussion}

### 2014-05-08 docs/paper.tex
final

@@
 Many approaches augment the agents with such features as working memory \cite{winder2012role,hu2003particle}.
-Our paper builds upon the work of Rodriguez & Reggia in \cite{rodriguez2004extending}, who added both working memory and a finite state machine (FSM) controller that switched the agent between different sets of movement behaviors according to its local environment and current goal.
+Our paper builds upon the work of Rodriguez \& Reggia in \cite{rodriguez2004extending}, who added both working memory and a finite state machine (FSM) controller that switched the agent between different sets of movement behaviors according to its local environment and current goal.
 They were able to demonstrate that a team of such agents was able to solve a resource locate-and-collect problem, and that a team programmed with flocking behaviors outperformed teams of agents that did not influence each other's movements.
@@
 \section{Discussion}
-The evolution designed particles surpassed their human design counterparts through a qualitatively novel mechanism; the human designed team prefered to stay in one or two larger flocks; whereas the high performing evolved teams split into more, smaller flocks and were able to exploit the search space more effectively.
+A nature-inspired, evolutionary technique was successfully able to design particles that not only exhibited collective, emergent behavior, but also surpassed their human design counterparts through a qualitatively novel mechanism.
+From epirical observation of the best performing teams, it appears that the human designed team prefered to stay in one or two larger flocks; whereas the high performing evolved teams split into more, smaller flocks and were able to exploit the search space more effectively.
+Because of the limited time frame in the simulation; this was clearly a better strategy, although additional experiments would be needed to determine the success of these particles over a larger time period.
 
-This work described herein represents only the first steps in the application of evolutionary techniques to this problem domain.
-By mutating and applying selective pressure to the movement component parameters, we were able to show that a machine could optimize the parameters better than a person.
-This was a form of optimization, but it did not result in novel behaviors.
-Indeed, since we hard-wired the transitions between the states (allowing for the guarding threshold to evolve) and allowed the evolver to mutate existing movement components in a given state but not to add them (i.e., we pre-defined which components could have non-zero weight for each state), we held constant the "meaning" of each state.
+The computational requirements for performing such an evolution were significant particularly in terms of time.
+However, parallel computation strategies enabled us to gather initial results.
+Future work might include MapReduce implementations of the evolutionary computation to distribute tasks across a cluster of computers, as well as longer runs to ensure more significant evolutionary behavior.
+In order to carry out further experimentation at scale, some larger computational framework would be needed to implement the demands of the simulations.
 
-The next step would be to evolve the structure of the FSM itself; not only to allow the evolver to mix-in any component to any state, but to allow it to add new states, delete old states, and change the transitions between the states.
+In particular, the work described in this paper represents only the first steps in the application of evolutionary techniques to the design of particles that exhibit flocking behavior.
+By mutating and applying selective pressure to the movement component parameters, we were able to show that a machine could optimize bottom-up velocity parameters better than a person, along with some limited state transition logic.
+However, because the states of the FSM controller were encoded in the simulation, we have not shown the ability to generate novel logical behaviors; simply better collective strategies.
+
+The next steps for this research would be to continue to evolve the structure of the FSM in more meaningful ways.
+In particular, new states should be generated by the configuration of their individual velocity components as well as other evolved conditional rules; not just the state parameters.
+Of interest is the opportunity to evolve the transitions between states in a more significant manner, and then to go further and evolve the size and structure of the FSM by adding and removing states.
+
+One essential research question that should be answered is whether or not some form of cross-over is possible in the FSM networks of particle swarms.
+If so, this will open up more opportunties for searching the optimal FSM space for collective movments.
+Other research questions might include better fitness functions that utilize the state of the entire world (for example, punishing swarms for allowing enemy agents to collect resources), and speciation for different strategies of evolution.
 

### 2014-05-09 bin/h2h.py
head to head evaluation

@@ -1,0 +1,88 @@
+#!/usr/bin/env python
+
+# h2h.py
+# Head to head tasking to compute performance against human design
+#
+# Author:   Benjamin Bengfort <benjamin@bengfort.com>
+# Created:  Fri May 09 16:10:50 2014 -0400
+#
+# Copyright (C) 2014 Bengfort.com
+# For license information, see LICENSE.txt
+#
+# ID: h2h.py [] benjamin@bengfort.com $
+
+"""
+Head to head tasking to compute performance against human design
+"""
+
+##########################################################################
+## Imports
+##########################################################################
+
+import os
+import sys
+import argparse
+
+from evolve.tasks import head2head
+
+##########################################################################
+## Command Line Description
+##########################################################################
+
+DESCRIPTION = "Head to head tasking to compute performance against human design"
+EPILOG      = "(C) Copyright University of Maryland 2014"
+VERSION     = "1.0"
+
+##########################################################################
+## Worker functions
+##########################################################################
+
+def queue_evaluation(args):
+    """
+    Queues several evaluations to be run simultanenously
+    """
+
+    config = args.config[0]
+    outdir = os.path.abspath(os.path.expandvars(os.path.expanduser(args.outdir)))
+
+    if os.path.exists(outdir):
+        if not os.path.isdir(outdir):
+            return "Output path '%s' is not a directory" % outdir
+        if len(list(name for name in os.listdir(outdir) if name.startswith(args.prefix))) > 0:
+            return "Output path '%s' is not empty!" % outdir
+    else:
+        os.makedirs(outdir)
+
+    for idx in xrange(1, args.trials+1):
+        outpath = os.path.join(outdir, '%s_%02i.csv' % (args.prefix, idx))
+        head2head.delay(config, outpath, args.iterations)
+
+    return '%i tasks queued. Evaluating on %i iterations.\nResults written to: %s' % (args.trials, args.iterations, outpath)
+
+##########################################################################
+## Main Method
+##########################################################################
+
+def main(*args):
+
+    # Construct the argument parser
+    parser = argparse.ArgumentParser(description=DESCRIPTION, epilog=EPILOG, version=VERSION)
+    subparsers = parser.add_subparsers(title='commands', description='Administrative commands for evaluating simulations.')
+
+    trial_parser = subparsers.add_parser('evaluate', help='Queue a number of evaluations (trials) to evaluate design performance.')
[…]

### 2014-05-09 evolve/tasks.py

@@ -17,6 +17,7 @@ The simulation tasks required for Celery
 ## Imports
 ##########################################################################
 
+import csv
 import time
 
 from swarm import World
@@ -29,6 +30,9 @@ from swarm.exceptions import SimulationException
 
 @app.task
 def runsim(configuration):
+    """
+    Run a simulation for the given number of timesteps and return fitness.
+    """
     start = time.time()
     world = World(ally_conf_path=configuration)
 
@@ -37,6 +41,40 @@ def runsim(configuration):
             world.update()
         except Exception as e:
             break
+
+    finit = time.time()
+    delta = finit - start
+
+    return {
+        'fitness':     world.ally_home.stash,
+        'run_time':    delta,
+        'iterations':  world.time,
+        'home_stash':  world.ally_home.stash,
+        'enemy_stash': world.enemy_home.stash,
+    }
+
+@app.task
+def head2head(configuration, outpath, iterations=10000):
+    """
+    Run a head to head simulation using the configuration for the black
+    team against the red team. Write detailed stats out to the outpath.
+    Can also specificy the number of iterations to run the simulation for.
+    """
+    start = time.time()
+    world = World(ally_conf_path=configuration, maximum_time=iterations)
+
+    with open(outpath, 'w') as outfile:
+        writer = csv.writer(outfile)
+        header = ('black', 'red',) + ('deposit',) * world.deposits
+        writer.writerow(header)
+        writer.writerow(world.status())
+
+        for step in xrange(world.iterations):
+            try:
+                world.update()
+                writer.writerow(world.status())
+            except Exception as e:
+                break
 
     finit = time.time()
     delta = finit - start

### 2014-05-09 swarm/world.py

@@ -126,6 +126,7 @@ class World(object):
 
         # Initialize resources
         self.add_agents(initialize_resources())
+        self.resources = [agent for agent in self.agents if agent.idx.startswith('mineral')]
 
     def add_agent(self, agent):
         agent.world = self
@@ -141,3 +142,12 @@ class World(object):
         for agent in self.agents:
             agent.blit()
         self.time += 1
+
+    def status(self):
+        """
+        Reports the number of resources in the bases and resource depots
+        at the given timestamp. (Ordered tuple).
+        """
+        bases = (self.ally_home.stash, self.enemy_home.stash)
+        depot = tuple(depot.stash for depot in self.resources)
+        return bases + depot

### 2014-05-09 bin/h2h.py
minor change

@@ -57,7 +57,7 @@ def queue_evaluation(args):
         outpath = os.path.join(outdir, '%s_%02i.csv' % (args.prefix, idx))
         head2head.delay(config, outpath, args.iterations)
 
-    return '%i tasks queued. Evaluating on %i iterations.\nResults written to: %s' % (args.trials, args.iterations, outpath)
+    return '%i tasks queued. Evaluating on %i iterations.\nResults written to: %s' % (args.trials, args.iterations, outdir)
 
 ##########################################################################
 ## Main Method

### 2014-05-11 bin/h2h.py
evaluation of results

@@ -21,6 +21,7 @@ Head to head tasking to compute performance against human design
 
 import os
 import sys
+import csv
 import argparse
 
 from evolve.tasks import head2head
@@ -31,7 +32,7 @@ from evolve.tasks import head2head
 
 DESCRIPTION = "Head to head tasking to compute performance against human design"
 EPILOG      = "(C) Copyright University of Maryland 2014"
-VERSION     = "1.0"
+VERSION     = "1.1"
 
 ##########################################################################
 ## Worker functions
@@ -59,6 +60,58 @@ def queue_evaluation(args):
 
     return '%i tasks queued. Evaluating on %i iterations.\nResults written to: %s' % (args.trials, args.iterations, outdir)
 
+def consolidate(args):
+    """
+    Aggregates the trial into a cohesive result
+    """
+
+    results = []
+    files   = 0
+
+    for name in os.listdir(args.results[0]):
+        # Filtering
+        if not name.startswith(args.prefix): continue
+
+        # File handling
+        files += 1
+        path  = os.path.join(args.results[0], name)
+
+        with open(path, 'r') as data:
+            reader = csv.reader(data)
+            reader.next() # Skip the first line
+            for idx, row in enumerate(reader):
+                if len(results) == idx:
+                    results.append(list(float(v) for v in row))
+                else:
+                    for jdx, val in enumerate(row):
+                        results[idx][jdx] += float(val)
+
+    averages = []
+    for row in results:
+        averages.append(tuple(val/files for val in row))
+
+    writer = csv.writer(args.outpath)
+    for row in averages:
+        writer.writerow(row)
+
+    if args.graph:
+        try:
+            import matplotlib.pyplot as plt
+        except ImportError:
+            print "matplotlib is required to graph results"
+
+        fig, axe = plt.subplots(figsize=(10,7))
+
+        axe.plot([r[0] for r in averages], 'k-')
+        axe.plot([r[1] for r in averages], 'r-')
+
+        axe.legend(('black team', 'red team'), 'upper left')
+        axe.set_title('Mean Resources Collected over Time')
+
+        plt.show()
+
+    return "Aggregated results from %i files." % (files)
+
 ##########################################################################
 ## Main Method
 ##########################################################################
@@ -76,6 +129,13 @@ def main(*args):
     trial_parser.add_argument('-i', '--iterations', type=int, default=10000, help='Number of iterations to evaluate on.')
[…]

### 2014-05-11 docs/addendum/main.tex
addendum writeup

\documentclass[letter]{article}

\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[top=1.5in, bottom=1.5in, left=1.5in, right=1.5in]{geometry}

\title{Results Addendum: Evolutionary Design of Particle Swarms}

\author{Benjamin Bengfort, Kevin Harrison, Phillip Kim}

\date{\today}

\begin{document}
\maketitle

Our second experiment consisted of changing the evolutionary computation mechanism slightly to make the search space more consistent. Instead of including one elite in the next generation, we carried five elites over. We also revised our mutation operator code to ensure that it was mutating in small, linear steps. Finally, we implemented recombination to try to allow genetic material from elites to influence the entire population.

Our second experiment used 50 individuals per population and ran for 100 generations. With the revised code, you can see in Figure \ref{fig:meanfit} that our experiment converged after approximately 50 generations. The best individual was discovered in population 97, but the maximum fitness individuals were also very close and consistent after 50 generations. This was a much better result than our first evolutionary computation experiment which had extremely variable mean fitness and the each generation exhibited much different behavior than their parent generation.

\begin{figure}
	\centering
	    \includegraphics[width=0.75\textwidth]{figures/meanfit_experiment2}
    \caption{\label{fig:meanfit}Fitness per generation}
\end{figure}

The best evolved particle not only exhibited swarm behavior, but performed significantly better at the locate and collect task than the human designed particle. We compared the performance of both the evolved and human designed particles vs. a human designed particle by running each simulation 32 times then computing the mean resources in the home base of each team at every timestep. Further, we extended the maximum timesteps for the simulation from 10,000 (used to determine fitness) to 20,000 to see if any different behavior resulted and to ensure that the swarms had enough time to collect all the resources.

As illustrated in Table \ref{table:meanfitness}, the evolved swarm collected over 100 more resources on average than the human designed swarm, with a significantly lower standard deviation. Averaged across all timesteps, Figure \ref{fig:head2head} shows the performance of the human designed vs human designed and evolution designed vs human designed swarms. Competing against each other, the human design swarms do equivalently in competition. However the evolution designed swarm is able to not only capture more resources faster, but around timestep 7000, begins to steal resources from the enemy base on average.

Qualitatively, the evolved swarm exhibits a very different spreading behavior than the human designed swarm. As exhibited in Figure \ref{fig:spreading} the evolved particles spread more radially, able to explore more of the world earlier on. Another characteriation of this is the size of the swarm, which is much larger, but then breaks apart into smaller swarms once resources are discovered, in order to better exploit more resources deposits at once. In contrast, the human designed particles tend to stay in one or two larger flocks and only exploit one depot at a time.

\begin{table}
    \renewcommand{\arraystretch}{1.5}
    \centering
    \begin{tabular}{l | c c cc}
        \hline
        Simulation & Mean & Std Dev & Max & Min \\
        \hline
        human vs human & 249.125 & 42.618 & 319 & 177 \\
        evolved vs human & 397 & 4.833 & 400 & 386 \\
        \hline
    \end{tabular}
    \caption{Fitness of particles competing against human designed particle.}
    \label{table:meanfitness}
\end{table}

\begin{figure}
\centering
\begin{minipage}{0.5\textwidth}
	\centering
    	\includegraphics[width=\textwidth]{figures/designed_v_designed}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
	\centering
    	\includegraphics[width=\textwidth]{figures/evolved_v_designed}
\end{minipage}
	\caption{\label{fig:head2head}Evolution Designed vs Human Designed Swarms}
\end{figure}

\begin{figure}
	\centering
    	\includegraphics[width=0.75\textwidth]{figures/spreading}
    \caption{\label{fig:spreading}Spreading behavior of two types of swarms.}
\end{figure}

\end{document}
